---
title: "Flux Course 2026: Carbon cycle model-data fusion"
author: "Michael Dietze"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(compiler)
library(tidyverse)
library(mvtnorm)
source("R/functions.R")
source("R/utils.R")

## configuration

#### SET THE ENSEMBLE SIZE
ne = 50 ## should be 200 - 5000, depending on what your computer can handle

timestep = 1800 #seconds

start_date = as.Date("2015-07-01")

outdir = "./" ## output directory for saving files
```

# Super Simple Ecosystem Model

Let's begin by defining our model itself, as well as a number of ancillary 
functions that will be useful in simulation and analysis. The model below is 
very simple but is complex enough to have some chance at capturing observed 
variability. FYI the code for this model, and most of the custom functions below, 
live in R/functions.R.

```{r}
SSEM.orig
```

In terms of inputs, the model has three state variables (X) that are all expressed in
terms of carbon (Mg/ha): Leaf Biomass (X[,1]), Structural Biomass (wood, roots, etc; X[,2]),
and soil organic matter (SOM; X[,3]). The modeled variables are thus the columns of X, while (as we'll see shortly) the different rows represent different replicate ensemble members. To be able to run the model we'll need to define the starting values of these variables (see **Initial Conditions**)

The model also has only two drivers, photosynthetically active radiation (PAR) and air temperature. We pass these into the model as a data frame "inputs" with columns "PAR" and temp" (see **Drivers**). The model parameters are similarly passed in through the data frame "params" and are defined in the **Parameters** section below.

Within the model we first 
estimate LAI from Leaf Biomass and specific leaf area (SLA). Using LAI and light we estimate GPP 
using a simple light use efficiency approach. GPP is then allocated to 
autotrophic respiration (Ra), leaf NPP, and woody NPP based on fixed allocation fractions. These leaf and wood 
biomass pools can then turns over into SOM as litterfall and Coarse Woody Debris / mortality assuming fixed turnover rates. 
SOM is then lost via heterotrophic respiration, which is assumed to be proportional to the size of the SOM pool and to follow a standard Q10 temperature sensitivity.  



# Initial Conditions

Having defined our model, the next step is to define the ensemble size and 
generate an ensemble estimate of the initial state variables. To do so we'll use
the estimates that are reported in the Ameriflux BADM Meta-data files for the 
site. Since we're only relying on two different estimates of pool size to 
calculate our mean and standard deviation, and neither estimate has a reported 
error, these should be taken as "demonstration only" rather than as 
"Best practices". In a real application one would want to account for the 
sampling error associated with the number of vegetation plots or soil cores 
measured, the measurement error in the soil C and tree DBH, and the allometric 
uncertainty in converting from DBH to leaf and stem biomass. In other words, our
pool sizes are likely a lot less certain than what we take them to be in this 
exercise.

There are a few assumptions in developing the initial conditions
- The mean and standard deviation are only calculated using two observations. 
  As stated above, this is not best practice.
- The wood state is the combination of stems, coarse roots, and fine roots
- The SOM (soil) state is the combination of litter, downed coarse woody debris,
  and soil.
- All values are converted from g/m2 to Mg/ha

```{r}
### Initial State (Mg/ha)
### These specific data points were extracted from the US-NR1 Ameriflux BADM files
### https://ameriflux.lbl.gov/sites/siteinfo/US-NR1
### METADATA: https://ameriflux.lbl.gov/data/badm/badm-standards/ 
### data is also available at https://bit.ly/3POkhq9
#library(readxl)
#badm = readxl::read_xlsx(file.path("AMF_AA-Flx_FLUXNET-BIF_CCBY4_20220606.xlsx")) %>% filter(SITE_ID == "US-NR1")
#write_csv(x=badm,file="data/US-NR1_BADM.csv")
#Bwood = badm %>% filter(VARIABLE == "AG_BIOMASS_TREE") %>% select(DATAVALUE) %>% type_convert() ## unsuccessful crack at automating extraction
Bwood = 14500 * 1e-6 * 10000 ## convert g/m2 -> Mg/ha
Bleaf = 2950 * 0.01 * 4.2/14 ## last term is a correction factor as raw obs generates a LAI that is too high
SOM = c(1.57, 1.58) + c(0.49, 1.39) + c(2.06, 2.59) * 1e-3 * 10000 ## sum up litter, CWD, and soil; change units (US-ME2)
X = as.matrix(c(mean(Bleaf), mean(Bwood), mean(SOM)))
### sample initial condition ensemble members
if(ne > 1){
  X = as.matrix(cbind(
      rnorm(ne, X[1], Bleaf*0.1), ## no variability in data, assume 10%
      rnorm(ne, X[2], Bwood*0.1), ## no variability in data, assume 10%
      rnorm(ne, X[3], sd(SOM))))
}
X.orig = X ## make a copy so that we can run different experiments

## visualize initial condition priors
pool.lab = c("leaf", "wood", "SOM")
for(i in 1:3){
  hist(X[, i], main = pool.lab[i])
}
```

# Drivers
```{r}
## data downloaded from FLUXNET2015 "fullset" product, subset to 2015 for this analysis
# flux <- read_csv("AMF_US-NR1_FLUXNET_FULLSET_HH_1998-2016_3-5.csv")
# flux <- flux %>% filter(date > as.Date("2015-01-01"))
# write_csv(x=flux,file= "data/US-NR1_Flux2015.csv")
flux <- read_csv("data/US-NR1_Flux2015.csv")
date = strptime(flux$TIMESTAMP_START,format="%Y%m%d%H%M")

inputs <- data.frame(
  date = date,
  temp = flux$TA_ERA, ## gap filled air temperature (Celsius)
  PAR  = flux$SW_IN_ERA / 0.486 ## gap filled PAR, conversion From Campbell and Norman p151
)
plot(inputs$date,inputs$PAR,type='l')
plot(inputs$date,inputs$temp,type='l')
```

# Parameters: Priors & Direct constraints (trait data)

This section defines our initial _a priori_ estimates of the model parameters before we confront the model with the observed eddy covariance data. As you see below, these are defined through a combination of site-specific trait data, literature values, and expert option. In all cases we aimed to define not just a single "default" value for the parameters but also our _a priori_ uncertainty. This prior uncertainty is important for understanding model sensitivity/uncertainty and for constraining the model calibration. For example, if a particular parameter is well known based on trait data, we wouldn't want to give it as much "wiggle room" as parameters that are based on rough guesses.
```{r}
## ancillary data from Ameriflux BADM metadata
SLA = 1e3/c(193.7,205.1,237.7)     ## m2/kg
litterfall = c(215.8)*0.01*3 ## gC/m2/yr->Mg/ha/yr

### initial params
params = list()

## univariate priors: expert opinion
params$alpha = rlnorm(ne, log(0.02), 0.05)     ## light use efficiency
params$Q10 = rnorm(ne, 2.1, 0.1)               ## soil respiration Q10
params$Rbasal = rlnorm(ne, log(0.2), 1) / (params$Q10^2.5) ## Soil basal respiration (umol/m2/sec per Mg/ha of SOM)

## multivariate prior on allocation parameters
## assume that NPP is ~50% of GPP on average (Litton et al 2007)
Ra = 0.5                                    
## prior mean on allocation, assume leaf NPP is 31.5% of total (Quaife et al 2008)
alloc = matrix(c(Ra, (1 - 0.315) * (1 - Ra), 0.315 * (1 - Ra)), 1) 
## draw effective sample size to add stochasticity to prior
Neff = matrix(rpois(ne, 100), ne)
rdirichlet <- cmpfun(rdirichlet.orig)         ## byte compile to speed up
## prior on fractional allocation to [Ra, wood, leaf] (each draw must sum to 1)
params$falloc = rdirichlet(ne,Neff%*%alloc) 

## Specific leaf area
params$SLA = rnorm(ne, mean(SLA), sd(SLA)) 

## simulate litterfall turnover based on observed 
## litterfall rate and Bleaf prior (initial condition)
if(length(litterfall)>1){
  sdlitterfall = sd(litterfall)
} else {
  sdlitterfall = 0.1*litterfall ## wild assumption
}
lit = rnorm(10000,mean(litterfall),sdlitterfall/sqrt(2))/     
  rnorm(10000,mean(X[,1]),sd(X[,1])/sqrt(2))     ## X1 = leaf biomass 
## draw prior mean and sd; convert turnover per year -> turnover per timestep
lit.mu = rnorm(ne,mean(lit),sd(lit))*timestep/86400/365  
lit.sd = 1/sqrt(rgamma(ne,10,10*var(lit)))*timestep/86400/365
litterfall.param = beta.match(lit.mu,lit.sd^2)
## match moments and draw litterfall prior
params$litterfall = rbeta(ne,litterfall.param$a,litterfall.param$b) 

## draw prior mean based on background tree mortality rate of 1/142 per year (Dietze et al 2011)
mortality.mu = 1/rpois(ne,142)*timestep/86400/365             
## draw prior sd assuming a 50% CV
mortality.sd = rbeta(ne,4,4)*mortality.mu*timestep/86400/365         
## match moments and draw mortality prior
mortality.param = beta.match(mortality.mu,mortality.sd^2)
params$mortality = rbeta(ne,mortality.param$a,mortality.param$b)  

## flatten to a data frame
params = as.data.frame(params)
```

### plot parameter values
```{r, fig.asp=1}
plot_params(params)
```

# Prior Predictive Checks: Monte Carlo Error Propagation


*Run model forward*
```{r}
system.time({
  output.ensemble = SSEM2(X=X.orig,
                          params = params,
                          inputs = inputs[,2:3], ## don't feed date col'n to model 
                          verbose= TRUE) 
})
```

*Plot timeseries*
Here we'll visualize both the 95% interval estimate, the median (black) and a random sample of 5 ensemble members (red) as examples
```{r}
plot_forecast(output.ensemble,sample=5)
```

If you like Ensemble Forecasting, you may also like...

* Alternative approaches to error propagation
  * Analytical distributional transformations
  * Analytical Moments
  * Linear Tangent approximations
* Uncertainty partitioning
* Predictability analyses (e.g. error vs lead time)
* Dietze 2017 Prediction in ecology: a first-principles framework. Ecological Applications 27: 2048–2060. DOI: 10.1002/eap.1589

# Assessment

At the half-hourly or hourly resolution, the _QC variable indicates if the corresponding record is a measured value (_QC = 0) or the quality level of the gap-filling that was used for that record (_QC = 1, _QC = 3 worse quality).
```{r}
## model
ci = apply(output.ensemble[, ,6], 1, quantile, c(0.025, 0.5, 0.975)) ## forecast confidence interval
                                                                     ## variable 6 = NEP
## observations
nep = -flux$NEE_VUT_REF ## change sign convention
nep.qc = flux$NEE_VUT_REF_QC
nep.unc = flux$NEE_VUT_REF_JOINTUNC

## plot timeseries
ylim = range(c(range(ci),range(nep,na.rm=TRUE)))
plot(inputs$date,ci[2,],ylim=ylim,col=1,pch=18,cex=0.2)
ciEnvelope(inputs$date, ci[1, ], ci[3, ], col = col.alpha("lightGrey", 0.5))
points(inputs$date,nep,col=nep.qc+1,cex=0.02)
legend("topright",legend=c("model",paste("QC",0:3)),
       pch = c(18,rep(1,4)),
       col=  c(1,1:4))

## predicted/observed plot
plot(ci[2,],nep,col=nep.qc+1,cex=0.2)
abline(0,1,lty=2)
```

In addition to these simple pre-calibration checks, there are a wide range of additional checks one can do to visually assess model performance, including but not limited to: 

* plots averaging over specific timescales (e.g., mean diurnal curve, mean seasonal curve, daily or monthly timeseries, "fingerprint" heatmaps of day-of-year by time-of-day) 
* plots of functional responses (e.g., predicted and observed relationships with temp and PAR)

TODO: add example code or link to validation lab

Also, in terms of assessing overall model performance it is critical to look at as many pools and fluxes as possible, not just NEE. I recommend the following order of assessment, which can be helpful for isolating the specific processes that may be miscalibrated or misrepresented in the model

1. LAI - should primarily be sensitive to initial conditions and SLA
2. unstressed GPP - should primarily be sensitive to photosythetic parameters. If you get the right GPP with the wrong LAI, you got the wrong GPP
3. stressed GPP - checks stress responses (e.g., moisture deficit), usually done by filtering the observed data for specific covariate conditions
4. NPP, biomass, biomass increment - if GPP is correct, this primarily checks autotrophic respiration, but also allocation and turnover
5. leaf and root litter, CWD - if NPP is correct, this primarily checks turnover and secondarily allocation
6. SOC - should primarily be sensitive to initial conditions
7. Soil respiration - primarily checks Rh. Functional responses can be particularly valuable. If you get Rh right but SOC wrong, you got Rh wrong
8. NEE - puts everything together. In can be hard to assess model errors from NEE alone, but diurnal curves can help separate errors in Re vs GPP
9. Soil Moisture (SM)
10. LE - I often check this much earlier for sites that have a strong carbon/water coupling

If you liked model assessment, you might also like...

* Model selection (AIC, DIC, wAIC, Predictive Loss)
* Model averaging

# Quantitative skill scores

In additionally to visually assessing model performance it 


```{r}
## apply QAQC
qaqc = (nep.qc == 0)
E = ci[2,qaqc]           ## model = expected
O = nep[qaqc] ## observed

## Model vs obs regressions
NEE.pre.fit = lm(O ~ E)

## performance stats
stats = as.data.frame(matrix(NA,2,4))
colnames(stats) <- c("RMSE","Bias","cor","slope")
rownames(stats) <- c("pre","post")
stats["pre","RMSE"] = sqrt(mean((E-O)^2))
stats["pre",'Bias'] = mean(E-O)
stats["pre",'cor']  = cor(E,O)
stats["pre",'slope'] = coef(NEE.pre.fit)[2]
stats["pre","R-squared"]=summary(NEE.pre.fit)$r.squared
knitr::kable(stats)

## predicted-observed
plot(E,O,pch=".",xlab="model",ylab='observed',main='NEE (umol/m2/sec)')
abline(0,1,col=2,lwd=2)
abline(NEE.pre.fit,col=3,lwd=3,lty=2)
legend("bottomright",legend=c('obs','1:1','reg'),col=1:3,lwd=3)
```

If you liked model skill scores, you might also like...

* Taylor Diagrams
* Quantile-based model diagnostics, Continuous Rank Probability Score (CRPS), and log score; see [Validation lab](https://github.com/EcoForecast/EF_Activities/blob/d92d8a44ee9d654ccbee569d9be3674821402a5c/Exercise_11_ModelAssessment.Rmd#L105)

# Sensitivity

```{r}
sa.summary = nee.sensitivity(ns=10)
## output is an array organized by parameter name, parameter samples, and variable
## variables are in the following order: 
vname = c("param","mean","RMSE","Bias","cor")

## plot sensitivities and calc linear approx
sens = sens_plot(var=2)

## ranked table of elasticities (i.e., standardized sensitivities)
knitr::kable(sens[order(abs(sens),decreasing = TRUE)]) 
```

# Cost Functions & Likelihoods

* What is a Likelihood?
* How does it allow us to update model parameters?


```{r}
sens_plot(var=3,line=FALSE) ## RMSE (minimize)
sens_plot(var=4,line=FALSE) ## bias (want as close as possible to 0)
sens_plot(var=5,line=FALSE) ## correlation (maximize)
```

```{r}
theta_bar = t(apply(params, 2, mean))
theta_bar[10] = stats["pre","RMSE"]
theta_range = apply(params, 2, range) 
Xbar      = matrix(apply(X.orig,2,mean),nrow=1)

likelihood <- function(param,Obs){
  param = data.frame(t(param))
  colnames(param) = c(colnames(params),"sigma")
  print(param)
  Model = SSEM(X = Xbar,
               params = param[,1:9],
               inputs = inputs[,2:3]) 
  lnL = -sum(dnorm(Obs,Model[qaqc],param[1,10],log=TRUE))
  return(lnL)
}

#lnL at the current parameter mean values
L0 = likelihood(param=theta_bar,Obs=O)

```

### Actual calibration takes a long time to run...
```{r}
system.time({
  fit = optim(theta_bar,likelihood,method="L-BFGS-B",
              lower=theta_range[1,],upper=theta_range[2,],
              hessian=TRUE,Obs=O)
})
```
#    user   system  elapsed 
# 3013.518   17.341 4280.111 

If you liked model calibration, you may also like...

* Markov Chain Monte Carlo (MCMC)
  * JAGS, NIMBLE
  * R's BayesianTools
* Hamiltonian Monte Carlo & STAN
* State Space Models
* Kalman Filter
* Ensemble Kalman Filter
* Particle Filter
* Using uncertainty analysis (Sobol, Monte Carlo, etc) to reduce the number of parameters considered


# Post-calibration prediction

```{r,fig.asp=1}
## updated parameter sample
mu = fit$par
H = abs(diag((fit$hessian + t(fit$hessian))/2))
#sigma = solve(H, tol = 1e-25)
#sigma = MASS::ginv(-H)
sigma = diag(1/H)
params_new = as.data.frame(mvtnorm::rmvnorm(ne,mu,sigma))
colnames(params_new) = c(colnames(params),"sigma")
plot_params(params_new)
```


```{r}
calibrated.ensemble = SSEM(X = X.orig,
                           params = params_new,
                           inputs = inputs[,2:3], ## don't feed date col'n to model 
                           verbose= TRUE) 
```











Timeseries for all variables
```{r}
plot_forecast(calibrated.ensemble,sample=5)

## model
ci.new = apply(calibrated.ensemble[, ,6], 1, quantile, c(0.025, 0.5, 0.975)) ## forecast confidence interval
                                                                     ## variable 6 = NEP
## plot timeseries
ylim = range(c(range(ci.new),range(nep,na.rm=TRUE)))
plot(inputs$date,ci.new[2,],ylim=ylim,col=1,pch=18,cex=0.2)
ciEnvelope(inputs$date, ci.new[1, ], ci.new[3, ], col = col.alpha("lightGrey", 0.5))
points(inputs$date,nep,col=nep.qc+1,cex=0.02)
legend("topright",legend=c("model",paste("QC",0:3)),
       pch = c(18,rep(1,4)),
       col=  c(1,1:4))

## predicted/observed plot
plot(ci[2,],nep,col=nep.qc+1,cex=0.2)
abline(0,1,lty=2)

## smoothed daily time series
date = as_date(inputs$date)
daily.nep = tapply(nep,date,mean,na.rm=TRUE)

daily.calibrated = apply(calibrated.ensemble[,,6],2,
                         function(x){tapply(x,date,mean,na.rm=TRUE)})
daily.cal.ci = apply(daily.calibrated, 1, quantile, c(0.025, 0.5, 0.975))

daily.orig = apply(output.ensemble[,,6],2,
                         function(x){tapply(x,date,mean,na.rm=TRUE)})
daily.orig.ci = apply(daily.orig, 1, quantile, c(0.025, 0.5, 0.975))

ylim = range(c(range(daily.orig.ci),range(daily.cal.ci),range(daily.nep,na.rm=TRUE)))
doy  = sort(unique(date))
plot(doy,daily.orig.ci[2,],ylim=ylim,col=1,pch=18,cex=0.5)
ciEnvelope(doy, daily.orig.ci[1, ], daily.orig.ci[3, ], col = col.alpha("lightGrey", 0.5))
points(doy,daily.cal.ci[2,],col=4,pch=18,cex=0.5)
ciEnvelope(doy, daily.cal.ci[1, ], daily.cal.ci[3, ], col = col.alpha("lightBlue", 0.5))
points(doy,daily.nep,col=2,cex=0.5)
legend("topright",legend=c("data","orig","calib"),
       pch = c(18,rep(1,3)),
       col=  c(2,1,4))

## daily predicted/observed
plot(daily.orig.ci[2,],daily.nep,ylim=ylim,xlim=ylim)
points(daily.cal.ci[2,],daily.nep,col=4)
abline(0,1,col=2,lty=2)

## mean diurnal curve
tod = inputs$date - as.POSIXct(date)-5
nep.diurnal = tapply(nep,tod,mean)
time.of.day = as.numeric(names(nep.diurnal))
cal.diurnal = tapply(ci.new[2,],tod,mean)
orig.diurnal = tapply(ci[2,],tod,mean)

ylim=range(c(range(cal.diurnal),range(orig.diurnal),range(nep.diurnal,na.rm=TRUE)))
plot(time.of.day,nep.diurnal,col=2,ylim=ylim)
lines(time.of.day,orig.diurnal)
lines(time.of.day,cal.diurnal,col=4)
legend("topright",legend=c("data","orig","calib"),
       pch = c(18,rep(1,3)),
       col=  c(2,1,4))

```



# Next steps

* Additional sites: EFI provides met and targets for all 47 NEON tower sites. Extending beyond these is possible but required nontrivial time spent on data processing.
* Additional data constraints to the Likelihood: Things that work particularly well are sensor-based observations that become available frequently (e.g., MODIS LAI, SMAP, soil respiration)
* Modify/improve the model: Lots of assumptions in SSEM are known to be false!
* Longer calibration (do parameters vary by season? by year?)

