---
title: "Exercise 01 — Running the Super Simple Ecosystem Model"
subtitle: "Fluxcourse 2026 · Bloomington, Indiana"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r metadata, echo=TRUE}
# Reproducibility header: record when this document was knitted, which
# commit of the repository it came from, and the full R environment.
# This block runs automatically every time you knit and appears in the HTML.
cat("Date:", format(Sys.time(), "%Y-%m-%d %H:%M"), "\n")
cat("Commit:", system("git rev-parse --short HEAD", intern = TRUE), "\n")
sessionInfo()
```

---

## Introduction

In the lectures you heard how most land surface models (LSMs) share a common
conceptual skeleton: carbon enters the ecosystem through photosynthesis, gets
partitioned among plant tissues, and eventually returns to the atmosphere via
autotrophic and heterotrophic respiration. The Super Simple Ecosystem Model
(SSEM) implements exactly that skeleton in its most stripped-down form.
Incoming light drives a light-use-efficiency calculation of gross primary
production (GPP); a fixed fraction of GPP is respired immediately as
autotrophic respiration (Ra) and the remainder is allocated between leaf and
wood growth; dead organic matter accumulates in a soil pool and decomposes at a
rate set by air temperature and a Q10 temperature sensitivity. There is no
energy balance, no water cycle, no nitrogen limitation, no phenology trigger, no
multi-layer canopy. That radical simplicity is intentional: because every arrow
of carbon can be traced from a single equation to a plotted output, SSEM is
ideal for developing a rigorous intuition for what process-based models actually
compute — and for diagnosing what goes wrong when they fail.

In this exercise you will run SSEM at the Tharandt spruce forest in Saxony,
Germany (FLUXNET site DE-Tha) for one calendar year, using only measured
shortwave radiation and air temperature as meteorological drivers. You will
examine the seasonal structure of the model output — GPP, net ecosystem
production (NEP), and carbon pool trajectories — and begin the core practice of
model evaluation: asking whether the output is physically plausible, and
identifying what observations would be required to answer that question more
rigorously. You are not yet comparing against observations; that is Exercise 02.
The goal here is to build a clear mental picture of what SSEM simulates and
what it deliberately ignores.

---

## Setup {.tabset}

```{r setup, message=FALSE, warning=FALSE}
# ── Libraries ─────────────────────────────────────────────────────────────────
# tidyverse loads dplyr (data manipulation), ggplot2 (figures), readr (fast
# CSV loading), tidyr (reshaping), and several other packages in one call.
library(tidyverse)

# lubridate provides convenient date-time parsing and arithmetic.
# We use it to parse FLUXNET ISO 8601 timestamps and filter by calendar year.
library(lubridate)

# ── Source the model code ─────────────────────────────────────────────────────
# R/functions.R contains the full SSEM implementation:
#   SSEM()              — core model: one 30-minute carbon cycle step
#   ensemble_forecast() — runs SSEM forward through an entire driver time series
#                         NOTE: uses 'ne' from the global environment (see Chunk 3)
#   ParticleFilter()    — sequential data assimilation (used in Exercise 02)
#   plot_forecast()     — 12-panel output summary with uncertainty envelopes
#   plot_params()       — diagnostic density plots of parameter distributions
#   rdirichlet.orig()   — draws from a Dirichlet distribution; used to sample
#                         carbon allocation fractions that must sum to 1
#   beta.match()        — Beta distribution moment matching for turnover priors
#   flux.uncertainty()  — heteroskedastic flux error model (Exercise 02)
#   varnames, units     — character vectors naming the 12 model output variables
source("../R/functions.R")

# R/utils.R contains lower-level helper functions:
#   ciEnvelope()    — draws a shaded credible interval polygon on a base R plot
#   col.alpha()     — adds alpha transparency to a colour specification
#   wtd.quantile()  — quantile function that respects particle filter weights
#   solar_geom()    — computes potential top-of-atmosphere shortwave radiation
#   mat2mcmc.list() — converts parameter matrices to CODA mcmc.list objects
source("../R/utils.R")
```

---

## Load and inspect drivers

```{r drivers, message=FALSE, fig.width=10, fig.height=6,
    fig.cap="Half-hourly PAR and air temperature at DE-Tha during 2000. Each vertical spike in PAR is one day's sunlight; the seasonal envelope shows peak insolation in June–July. Temperature drops below zero during winter."}

# ── Read the DE-Tha site data ─────────────────────────────────────────────────
# DE-Tha is a managed Norway spruce plantation (Picea abies) in Tharandt,
# Saxony, Germany. It has been a continuous FLUXNET site since 1996, making
# it one of the longest eddy covariance records in Europe. The _HH suffix
# denotes the half-hourly (HH) product: one row per 30-minute period.
drivers_raw <- read_csv(
  "../data/DE-Tha/DE-Tha_HH.csv",
  show_col_types = FALSE
)

# FLUXNET uses -9999 as a sentinel for missing observations (not the same as
# a true zero). Replace these with NA so R treats them as missing data rather
# than including them in calculations.
drivers_raw <- drivers_raw |>
  mutate(across(where(is.numeric), \(x) replace(x, x == -9999, NA)))

# ── Filter to one calendar year ───────────────────────────────────────────────
# We use the year 2000: a meteorologically representative year for this site
# with no major disturbance events, giving a clean seasonal cycle to study.
# Year 2000 is a leap year (366 days × 48 half-hour periods = 17,568 rows).
# After you have worked through this exercise, try a different year and
# compare the shape of the model output — interannual variability is real and
# the model will reflect the variability in the drivers.
drivers <- drivers_raw |>
  mutate(
    # Parse the ISO 8601 timestamp. The trailing "Z" indicates UTC;
    # ymd_hms() handles this automatically.
    datetime = ymd_hms(DATETIME_START)
  ) |>
  filter(year(datetime) == 2000) |>
  select(datetime, SW_IN_F, TA_F)

# Confirm the expected row count. A leap year has 366 × 48 = 17,568 rows.
cat("Timesteps in 2000:", nrow(drivers), "(expect 17,568 for a leap year)\n")

# ── Convert to model input units ──────────────────────────────────────────────
# SSEM's photosynthesis routine expects exactly two driver columns:
#   PAR  — photosynthetically active radiation (μmol photon m⁻² s⁻¹)
#   temp — air temperature (°C)
#
# SW_IN_F is gap-filled incoming shortwave radiation in W m⁻².
# PAR occupies the 400–700 nm waveband, roughly 45% of total shortwave.
# Within that waveband, 1 W m⁻² ≈ 4.6 μmol photons.
# Combined: PAR (μmol m⁻² s⁻¹) ≈ SW_IN_F (W m⁻²) × 0.45 × 4.6 ≈ SW_IN_F × 2.1
# We floor at 0 because sensors occasionally report tiny negative values at night
# due to calibration drift; the model does not allow negative PAR.
#
# TA_F is gap-filled air temperature already in °C — no unit conversion needed.
drivers <- drivers |>
  mutate(
    PAR  = pmax(SW_IN_F * 2.1, 0),
    temp = TA_F
  )

# ┌─────────────────────────────────────────────────────────────────────────────┐
# │  TIMESTEP NOTE — important if you switch sites                              │
# │                                                                             │
# │  DE-Tha (this script) and DK-Sor both use the FLUXNET "_HH_" half-hourly  │
# │  product: one row every 30 minutes → 48 rows per calendar day.             │
# │                                                                             │
# │  US-MMS uses the AmeriFlux "_HR_" hourly product (see the note at the      │
# │  bottom of this chunk): one row per hour → 24 rows per calendar day.       │
# │                                                                             │
# │  SSEM itself has no fixed timestep requirement — it loops over whatever    │
# │  rows you supply. But the number of rows per day affects two things:       │
# │                                                                             │
# │  1. The x-axis of plot_forecast() counts timestep numbers, not dates.      │
# │     For DE-Tha, step 48 = end of day 1; for US-MMS it would be step 24.   │
# │                                                                             │
# │  2. In Exercise 02, converting model output from μmol m⁻² s⁻¹ (a rate)   │
# │     to gC m⁻² day⁻¹ (a daily total) requires multiplying by the step      │
# │     length in seconds: 1800 s for 30-min sites, 3600 s for hourly sites.  │
# │     Using the wrong value introduces a factor-of-two error in daily totals │
# │     for US-MMS — easy to miss, hard to diagnose in a plot.                │
# │                                                                             │
# │  If you switch to US-MMS or DK-Sor data: (a) replace the file path above, │
# │  (b) update the expected-row-count comment (17,568 → 8,784 for US-MMS in  │
# │  a leap year), and (c) pass the correct step length to the unit conversion │
# │  in Exercise 02 via the site_timestep list defined in Chunk 1 of that Rmd. │
# └─────────────────────────────────────────────────────────────────────────────┘

# ── What does SSEM ignore? ────────────────────────────────────────────────────
# PAR and temperature are sufficient to drive SSEM's two central equations:
#   GPP = α × (1 − exp(−0.5 × LAI)) × PAR       [light use efficiency model]
#   Rh  = Rbasal × BSOM × Q10^(temp / 10)        [Q10 soil decomposition model]
#
# A full LSM would also require:
#   • Precipitation / soil moisture  — drought limits stomatal conductance and
#     slows decomposition; SSEM cannot simulate any water stress.
#   • Vapour pressure deficit (VPD)  — high VPD closes stomata, reducing GPP;
#     SSEM's LUE is a fixed parameter, not a VPD-responsive one.
#   • Leaf nitrogen / Vcmax          — N availability sets the photosynthetic
#     capacity; SSEM lumps this entirely into the α parameter.
#   • Phenology (budburst, leaf-off) — SSEM keeps Bleaf roughly constant;
#     there is no spring green-up or autumn senescence signal.
# You will see the consequences of these omissions when you compare the output
# against observations in Exercise 02.

# ── Plot the drivers ──────────────────────────────────────────────────────────
# Plotting the raw half-hourly data (rather than daily means) shows the diurnal
# cycle embedded within the seasonal trend — both matter for model behaviour
# because SSEM calculates GPP every 30 minutes.
drivers |>
  select(datetime, PAR, temp) |>
  pivot_longer(
    cols      = c(PAR, temp),
    names_to  = "variable",
    values_to = "value"
  ) |>
  mutate(
    # Create human-readable labels for the facet panels.
    label = case_when(
      variable == "PAR"  ~ "PAR (μmol photon m⁻² s⁻¹)",
      variable == "temp" ~ "Air temperature (°C)"
    )
  ) |>
  ggplot(aes(x = datetime, y = value)) +
    geom_line(linewidth = 0.1, alpha = 0.35, colour = "grey20") +
    # Free y-axis scales because PAR and temperature are on completely
    # different scales; a shared axis would compress one variable unreadably.
    facet_wrap(~ label, ncol = 1, scales = "free_y") +
    labs(
      title   = "Meteorological drivers — DE-Tha 2000",
      subtitle = "Half-hourly; PAR converted from SW_IN_F (W m⁻²) × 2.1",
      x        = NULL,
      y        = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(strip.background = element_blank(),
          strip.text        = element_text(face = "bold"))
```

---

## Parameters and initial conditions

```{r parameters}
# ── Ensemble size ─────────────────────────────────────────────────────────────
# ensemble_forecast() reads 'ne' from the global environment — it must be
# defined before you call the function. A larger ne samples parameter
# uncertainty more thoroughly but increases run time in direct proportion.
# 100 members gives a stable uncertainty envelope for exploratory work.
ne <- 100

# Fix the random seed so your results match this document exactly.
# Change the seed to explore a different realisation of the ensemble.
set.seed(2026)

# ── Initial carbon pool sizes (X) ─────────────────────────────────────────────
# SSEM tracks three carbon pools, all in Mg C ha⁻¹:
#   column 1 — Bleaf:  leaf carbon
#   column 2 — Bwood:  wood (stem + coarse root) carbon
#   column 3 — BSOM:   soil organic carbon
#
# Every row of X is one ensemble member. All members start from the same
# point here because we are expressing our prior uncertainty entirely through
# the parameter distributions below. A spin-up run (running the model to
# equilibrium from arbitrary initial conditions) would give more defensible
# starting values; for this exercise we use literature-based estimates for a
# mature Norway spruce stand.
X <- matrix(
  data = c(
      3.0,   # Bleaf: spruce needle carbon. Literature values for mature Picea
             #   abies stands: 2–5 Mg C ha⁻¹. Needles live ~4 years; at any
             #   moment roughly four cohorts are present.
    100.0,   # Bwood: stem + coarse root carbon. Mature managed spruce in
             #   central Europe: 80–150 Mg C ha⁻¹ (varies with age and site).
    100.0    # BSOM: soil organic carbon to 1 m depth. Temperate forest mineral
             #   soils typically store 50–150 Mg C ha⁻¹; organic horizons under
             #   spruce can push this higher.
  ),
  nrow  = ne,
  ncol  = 3,
  byrow = FALSE   # each column is one pool; rows are ensemble members
)

# ── Parameter ensemble ────────────────────────────────────────────────────────
# We draw ne realisations of each parameter from a probability distribution
# that represents our prior uncertainty about the true value at this site.
# The spread propagates through the model and appears as the grey uncertainty
# envelope in Chunk 4. Wider priors → wider envelopes → more honest
# representation of what we do not know.
params <- data.frame(

  # α (alpha): canopy light use efficiency
  #   The slope of the GPP–PAR relationship: GPP = α × Beer's law × PAR
  #   Units: μmol CO₂ per μmol photon absorbed (dimensionless ratio)
  #   In a full LSM: derived from leaf-level Vcmax and Jmax through the
  #   Farquhar model; here compressed into one empirical constant.
  #   Norway spruce literature: 0.01–0.03 μmol μmol⁻¹ (Medlyn et al. 2003).
  alpha = rnorm(ne, mean = 0.02, sd = 0.003) |> pmax(0),

  # SLA: specific leaf area
  #   Converts leaf carbon mass to leaf area: LAI = Bleaf × SLA × 0.1
  #   (the 0.1 converts Mg ha⁻¹ to kg m⁻² for unit consistency).
  #   Units: m² leaf area per kg leaf dry mass
  #   In a full LSM: read from trait databases (TRY); Norway spruce: 4–12 m² kg⁻¹.
  SLA = rnorm(ne, mean = 8.0, sd = 1.2) |> pmax(1),

  # Rbasal: basal heterotrophic respiration rate
  #   Rh = Rbasal × BSOM × Q10^(temp / 10)
  #   Units: μmol CO₂ m⁻² s⁻¹ per Mg C ha⁻¹ of soil organic carbon.
  #   Interpretation: with BSOM = 100 Mg C ha⁻¹ and temp = 10 °C,
  #   Rbasal = 0.015 gives Rh ≈ 3 μmol m⁻² s⁻¹ — within the observed
  #   range for temperate forest soils (Janssens et al. 2001).
  #   In a full LSM: fitted from soil incubation data or inferred from the
  #   seasonal Rh–temperature relationship in flux tower data.
  Rbasal = rnorm(ne, mean = 0.015, sd = 0.003) |> pmax(0),

  # Q10: temperature sensitivity of heterotrophic respiration
  #   Each 10 °C increase in soil temperature multiplies Rh by Q10.
  #   Dimensionless; meta-analyses report 1.4–2.5; canonical value = 2.
  #   In a full LSM: fitted from soil warming experiments or from the
  #   slope of the seasonal Rh–temperature relationship at the tower.
  Q10 = rnorm(ne, mean = 2.0, sd = 0.3) |> pmax(1),

  # litterfall: fraction of leaf C transferred to BSOM per 30-min timestep
  #   Norway spruce needles live on average ~4 years, so ~25% of Bleaf
  #   falls to the soil each year: 0.25 / (365 × 48) ≈ 1.4 × 10⁻⁵ per step.
  #   Units: dimensionless fraction per 30-min timestep
  #   In a full LSM: output of a phenology model (budburst + senescence dates
  #   driven by photoperiod and temperature); SSEM treats it as constant.
  litterfall = rnorm(ne, mean = 1.4e-5, sd = 1.5e-6) |> pmax(0),

  # mortality: fraction of wood C transferred to BSOM per 30-min timestep
  #   Includes stem death and coarse root turnover. Managed spruce in Europe
  #   is harvested on ~80-year rotations → ~1.25% yr⁻¹ carbon loss from Bwood.
  #   Per 30-min step: 0.0125 / (365 × 48) ≈ 7.1 × 10⁻⁷
  #   Units: dimensionless fraction per 30-min timestep
  #   In a full LSM: output of a demographic or disturbance model.
  mortality = rnorm(ne, mean = 7.1e-7, sd = 7e-8) |> pmax(0),

  # sigma.leaf, sigma.stem, sigma.soil: process noise standard deviations
  #   Gaussian noise added to each pool at every timestep to represent model
  #   structural error — the cumulative mismatch between what the equations
  #   predict and what the real ecosystem does (missing processes, parameter
  #   error, numerical discretisation). Units: Mg C ha⁻¹.
  #   The particle filter (Exercise 02) can estimate these from the data.
  #   Leaf pool is small and tightly coupled to inputs, so gets smaller noise.
  sigma.leaf = rep(0.01, ne),
  sigma.stem = rep(0.10, ne),
  sigma.soil = rep(0.10, ne)
)

# ── Carbon allocation fractions ───────────────────────────────────────────────
# SSEM splits GPP into three competing fates; their fractions must sum to 1
# for every ensemble member. We draw from a Dirichlet distribution, which is
# the correct prior for a probability simplex (values in [0,1] summing to 1).
#
# Concentration parameters control the expected proportions:
#   falloc.1 → Ra (autotrophic respiration)  : 5 / 10 = 50% expected
#   falloc.2 → NPPwood (wood growth)         : 3 / 10 = 30% expected
#   falloc.3 → NPPleaf (leaf growth)         : 2 / 10 = 20% expected
#
# These are consistent with a carbon use efficiency (CUE = NPP/GPP) of ~0.50,
# which is close to the observed mean for temperate forests (Malhi 2012).
# Larger concentration values reduce the spread across ensemble members.
falloc <- rdirichlet.orig(ne, alpha = c(5, 3, 2))
params$falloc.1 <- falloc[, 1]   # fraction of GPP → Ra
params$falloc.2 <- falloc[, 2]   # fraction of GPP → wood NPP
params$falloc.3 <- falloc[, 3]   # fraction of GPP → leaf NPP

# Verify the constraint holds for every ensemble member.
stopifnot(
  all(abs(rowSums(params[, c("falloc.1", "falloc.2", "falloc.3")]) - 1) < 1e-10)
)

# Print the parameter ensemble summary so you can see the prior spread.
cat("\nParameter ensemble: mean ± SD across", ne, "members\n")
cat(sprintf("  %-12s  %10s  %10s\n", "parameter", "mean", "SD"))
cat(rep("-", 38), "\n", sep = "")
for (p in names(params)) {
  cat(sprintf("  %-12s  %10.3g  %10.3g\n", p, mean(params[[p]]), sd(params[[p]])))
}
```

---

## Run the model and examine output

```{r model_run, cache=TRUE, fig.width=12, fig.height=10,
    fig.cap="SSEM ensemble output for DE-Tha 2000. Grey envelope: 95% credible interval across 100 ensemble members. Dark line: ensemble median. x-axis: 30-minute timestep number (48 steps = 1 day; 17,568 steps = full leap year)."}

# ── Assemble inputs ───────────────────────────────────────────────────────────
# ensemble_forecast() expects a plain data.frame with exactly the columns
# 'PAR' and 'temp', one row per model timestep. The function loops over rows
# in chronological order, passing each row to SSEM() as the current drivers.
inputs <- drivers |>
  select(PAR, temp) |>
  as.data.frame()   # convert from tibble; some internal indexing expects data.frame

# Confirm dimensions match: inputs must have one row per timestep.
cat("Input rows:", nrow(inputs), " | Expected (leap year):", 17568, "\n")

# ── Run the ensemble ──────────────────────────────────────────────────────────
# ensemble_forecast() runs all ne ensemble members forward simultaneously.
# It reads 'ne' from the global environment (defined in the previous chunk).
# A counter prints every 336 steps (= 7 days); for a leap year expect ~52 lines.
# Run time is approximately 2–4 minutes on a modern laptop.
#
# The returned object is a 3-D array:
#   dim 1 — timestep          (one 30-min step per row)
#   dim 2 — ensemble member   (ne columns)
#   dim 3 — output variable   (12 variables; see list below)
#
# Variable index → name → units:
#    1  Bleaf      leaf carbon            Mg C ha⁻¹
#    2  Bwood      wood carbon            Mg C ha⁻¹
#    3  BSOM       soil organic carbon    Mg C ha⁻¹
#    4  LAI        leaf area index        m² m⁻²
#    5  GPP        gross primary prod.    μmol CO₂ m⁻² s⁻¹
#    6  NEP        net ecosystem prod.    μmol CO₂ m⁻² s⁻¹
#    7  Ra         autotrophic resp.      μmol CO₂ m⁻² s⁻¹
#    8  NPPw       wood net prim. prod.   μmol CO₂ m⁻² s⁻¹
#    9  NPPl       leaf net prim. prod.   μmol CO₂ m⁻² s⁻¹
#   10  Rh         heterotrophic resp.    μmol CO₂ m⁻² s⁻¹
#   11  litterfall leaf C to soil         Mg C ha⁻¹ per timestep
#   12  mortality  wood C to soil         Mg C ha⁻¹ per timestep
message("Starting model run — expect 2–4 minutes …")
output <- ensemble_forecast(X, params, inputs)

cat("\nOutput array dimensions [timesteps × ensemble members × variables]:\n")
print(dim(output))

# ── What does the ensemble spread represent? ──────────────────────────────────
# Each ensemble member carries a different parameter set drawn from the
# distributions in Chunk 3. When run forward, those differences compound over
# time into diverging pool trajectories and flux time series. The width of the
# 95% envelope at any timestep is therefore a direct measure of forecast
# uncertainty arising from parameter uncertainty. A wide envelope means the
# model output is sensitive to that source of uncertainty; a narrow envelope
# means the model is relatively insensitive to it. In Exercise 02 you will
# use eddy covariance NEP observations to weight ensemble members by
# how well they match the data — the particle filter — collapsing the
# envelope where observations provide information and leaving it wide where
# they do not.

# ── Plot all 12 output variables ──────────────────────────────────────────────
# plot_forecast() (from R/functions.R) computes the 2.5th, 50th, and 97.5th
# percentiles of the ensemble at each timestep and renders them as a shaded
# ribbon (ciEnvelope) around the median line. The function uses base R graphics
# rather than ggplot2 because it is from the upstream SSEM repository; we call
# it unchanged.
par(mfrow = c(4, 3),           # 4 rows × 3 columns = 12 panels for 12 variables
    mar   = c(3, 4, 3, 0.5))  # tighter margins so all panels fit comfortably
plot_forecast(output)

# Reset to single-panel layout for any subsequent plots.
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)

# ── Save output for Exercise 02 ───────────────────────────────────────────────
# Exercise 02 loads this RDS to compare SSEM output against observations.
# Saving here means you only need to run Exercise 01 once; re-knitting
# Exercise 02 will always find the file even without re-running the model.
saveRDS(output, "../data/ssem_detha_2000.rds")
message("Saved SSEM output → data/ssem_detha_2000.rds")

# ── Explore further ───────────────────────────────────────────────────────────
# Half-hourly driver data are available for two additional temperate forest sites:
#
#   data/DK-Sor/DK-Sor_HH.csv
#     Soroe, Denmark — European beech forest (Fagus sylvatica); deciduous,
#     so it has a very pronounced spring green-up and autumn leaf-off. Contrast
#     the seasonal GPP shape with DE-Tha.
#
#   data/US-MMS/  (daily and monthly only for this site; no HH file)
#     Morgan Monroe State Forest, Indiana — mixed deciduous broadleaf forest;
#     warmer summers and colder winters than Germany.
#
# To run SSEM at DK-Sor, replace DE-Tha_HH.csv with DK-Sor_HH.csv in Chunk 2
# and re-knit. Consider: does an evergreen model structure make sense for beech?
```

---

## Discussion questions

Work through the following questions before moving on to Exercise 02. There
are no single correct answers; the goal is to reason from what you know about
forest carbon cycling and what you can see in the model output panels above.

**1. Seasonal pattern of GPP.** Examine the GPP panel. Describe the seasonal
trajectory: when does GPP peak, when does it drop toward zero, and does the
shape match your expectation for a temperate Norway spruce forest? Spruce is
an evergreen conifer — unlike a beech or oak it retains its needles year-round
and can photosynthesize on warm winter days. Does the output reflect that
possibility? Think about what the light-use-efficiency formulation assumes
about stomatal behaviour in cold or dry conditions, and consider whether that
assumption is likely to exaggerate or suppress winter GPP at this site.

**2. Information needed to evaluate the output.** You have a model trajectory
but no observational reference to compare it against. List the data streams that
would allow you to assess whether this output is realistic, and for each one
note: (a) which model variable or flux it constrains most directly, (b) the
typical measurement uncertainty you would need to account for, and (c) any
mismatch in temporal or spatial scale between the observation and the model
output. Consider both eddy covariance fluxes available at DE-Tha through the
FLUXNET archive and independent pool-level measurements such as forest
inventory biomass, litter trap data, or soil carbon surveys.

**3. Ensemble spread and forecast uncertainty.** Compare the width of the grey
uncertainty envelope across the 12 panels. Which state variables or fluxes
show the widest spread relative to their magnitude, and which are tightly
constrained? What does this pattern tell you about which parameters dominate
forecast uncertainty? If you could make one additional field measurement to
reduce the uncertainty in the carbon-cycle forecast, which measurement would
you choose, and why?
