---
title: "Exercise 02 — Validating the Super Simple Ecosystem Model"
subtitle: "Fluxcourse 2026 · Bloomington, Indiana"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r metadata, echo=TRUE}
cat("Date:", format(Sys.time(), "%Y-%m-%d %H:%M"), "\n")
cat("Commit:", system("git rev-parse --short HEAD", intern = TRUE), "\n")
sessionInfo()
```

---

## Introduction

Running a model and validating a model are different activities. In Exercise 01
you ran SSEM and confirmed it produces plausible-looking output. That is a
necessary first step, but it is not validation: the model could produce any
seasonal cycle you chose simply by adjusting its parameters, and there is no
way to know whether its numbers are right without confronting them with
independent observations. Validation requires data that the model has not
already seen, measured at a spatial and temporal scale that actually matches
what the model simulates, and processed consistently enough that you are
comparing the same quantity. Each of those three conditions is harder to satisfy
than it sounds, and failing to notice a mismatch in any one of them is one of
the most common sources of spurious conclusions in model-data comparison studies.

This exercise takes you through that comparison work. You will bring together
four observational data streams — eddy covariance flux tower records from three
temperate forest sites (DE-Tha in Germany, DK-Sor in Denmark, US-MMS in
Indiana), FLUXCOM upscaled flux estimates, MODIS-derived LAI, and output from
one Earth system model (ESM) of your choice from the CMIP6 archive — and you
will compare all of them against each other and against SSEM. The sites span
two vegetation types: DE-Tha and DK-Sor are temperate European forests (a
Norway spruce plantation and a beech forest respectively), while US-MMS is a
mixed deciduous broadleaf forest in the Midwest. Together they expose the range
of carbon-cycle behaviour that a simple model like SSEM must explain — and
frequently cannot.

---

## Setup

> **Before you start:** This exercise loads the SSEM ensemble output array
> that was computed in Exercise 01. To make that file available, open
> `exercises/01_run_model.Rmd`, knit it once to completion, and then run the
> following line in the R console (or add it to the end of the `model_run`
> chunk and re-knit):
>
> ```r
> saveRDS(output, "../data/ssem_detha_2000.rds")
> ```
>
> Once that file exists on disk, this document is self-contained.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)   # dplyr, ggplot2, readr, tidyr, purrr, lubridate wrapper
library(lubridate)   # date arithmetic, ymd(), days_in_month()

# R/functions.R: SSEM model, ensemble_forecast, plot_forecast, rdirichlet.orig,
# flux.uncertainty, varnames/units character vectors, ParticleFilter
source("../R/functions.R")

# R/utils.R: ciEnvelope, col.alpha, wtd.quantile, solar_geom, mat2mcmc.list
source("../R/utils.R")

# ── Load SSEM ensemble output from Exercise 01 ────────────────────────────────
# ensemble_forecast() returns a 3-D array [timestep × ensemble × variable].
# We read it from disk so this document does not need to re-run the model.
rds_path <- "../data/ssem_detha_2000.rds"
if (file.exists(rds_path)) {
  ssem_output <- readRDS(rds_path)
  cat("SSEM output loaded:", paste(dim(ssem_output), collapse = " × "),
      "[timesteps × ensemble × variables]\n")
} else {
  stop(paste(
    "File not found:", rds_path, "\n\n",
    "Please knit 01_run_model.Rmd first and then run:\n",
    "  saveRDS(output, \"../data/ssem_detha_2000.rds\")\n",
    "in the console, then re-knit this document."
  ))
}

# ── Site timestep intervals ────────────────────────────────────────────────────
# Each flux tower site uses a different sub-daily measurement frequency.
# This list is the single source of truth used downstream in Chunk 3 to convert
# instantaneous flux rates (μmol CO₂ m⁻² s⁻¹) to per-timestep totals
# (gC m⁻² per step) and ultimately to daily totals (gC m⁻² day⁻¹).
#
# Getting the timestep wrong introduces a proportional error in every flux
# integral. For US-MMS vs DE-Tha, using the same constant for both would halve
# or double one site's daily totals — making two ecologically similar forests
# appear very different for a purely methodological reason.
#
# DE-Tha / DK-Sor: "_HH_" half-hourly FLUXNET product — 30 min intervals,
#   48 rows per day, 1800 seconds per timestep.
# US-MMS: "_HR_" hourly AmeriFlux product — 60 min intervals, 24 rows per day,
#   3600 seconds per timestep. The "_HR_" naming is an AmeriFlux convention that
#   differs from the legacy FLUXNET "_HH_" convention; see the comment in
#   Chunk 2 and data/US-MMS/extract_hr_workaround.R for the full explanation.
site_timestep <- list(
  "DE-Tha" = 30,   # minutes per timestep (half-hourly, _HH_ product)
  "DK-Sor" = 30,   # minutes per timestep (half-hourly, _HH_ product)
  "US-MMS" = 60    # minutes per timestep (hourly, _HR_ product)
)
```

---

## Load raw observations

> The code below loads every data source exactly as it arrives — no unit
> conversions yet. Read the column names and the raw values carefully. You are
> about to encounter one of the most routine but under-discussed problems in
> multi-source model evaluation: every system uses different units, different
> sign conventions, and different time resolutions. Before you can compare
> anything, you must understand what you have.

```{r load_raw, message=FALSE}
# ── Sub-daily FLUXNET data for all three sites ────────────────────────────────
# We load the finest available time resolution for each site.  The variable
# names differ between sites because DE-Tha and DK-Sor were processed through
# the fluxnet R package while US-MMS was extracted by a standalone workaround.
#
# DE-Tha_HH.csv — half-hourly (30 min), timestamp column: DATETIME_START
#   ISO 8601 format ("1996-01-01T00:00:00Z"); processed by fluxnet package.
#   Site: Norway spruce plantation, Tharandt, Germany; coverage 1996–2024.
#
# DK-Sor_HH.csv — half-hourly (30 min), timestamp column: DATETIME_START
#   Same format as DE-Tha; also processed by the fluxnet package.
#   Site: European beech forest, Sorø, Denmark; coverage 1996–2024.
#
# US-MMS_HR.csv — hourly (60 min), timestamp column: TIMESTAMP_START
#   Integer format YYYYMMDDHHMM (e.g., 199901010000 = midnight 1 Jan 1999).
#   NOT produced by the fluxnet package because flux_extract() v0.3.1 silently
#   skips files named "_HR_" (it searches for the legacy "_HH_" pattern from
#   older FLUXNET products). The file was extracted by the standalone workaround
#   at data/US-MMS/extract_hr_workaround.R — see that script for the full
#   explanation of the bug and the GitHub issue tracking the fix.
#   Site: mixed deciduous broadleaf forest, Morgan Monroe State Forest, Indiana;
#   coverage 1999–2023.
# col_select loads only the three columns that harmonise_subdaily() needs.
# The HH/HR files are 695 MB / 683 MB / 284 MB on disk and expand to ~2 GB each
# in memory when fully loaded; selecting only these columns reduces each to < 10 MB.
detha_hh_raw <- read_csv("../data/DE-Tha/DE-Tha_HH.csv", show_col_types = FALSE,
                          col_select = c(DATETIME_START, NEE_VUT_REF, NEE_VUT_REF_QC))
dksor_hh_raw <- read_csv("../data/DK-Sor/DK-Sor_HH.csv", show_col_types = FALSE,
                          col_select = c(DATETIME_START, NEE_VUT_REF, NEE_VUT_REF_QC))
usmms_hr_raw <- read_csv("../data/US-MMS/US-MMS_HR.csv",  show_col_types = FALSE,
                          col_select = c(TIMESTAMP_START, NEE_VUT_REF, NEE_VUT_REF_QC))

# FLUXNET/AmeriFlux fill value: -9999 means missing (not zero). Replace with NA.
# across(where(is.numeric)) catches every numeric column in one call.
replace_fluxnet_na <- function(df) {
  df |> mutate(across(where(is.numeric), \(x) replace(x, x == -9999, NA)))
}
detha_hh_raw <- replace_fluxnet_na(detha_hh_raw)
dksor_hh_raw <- replace_fluxnet_na(dksor_hh_raw)
usmms_hr_raw <- replace_fluxnet_na(usmms_hr_raw)

# ── US-MMS daily (DD) for CMIP6 Chunk 6 compatibility ────────────────────────
# Chunk 6 (CMIP6 comparison, monthly scale) accesses `usmms_raw` and expects a
# DATE column and daily-resolution GPP/RECO columns — the format produced by
# the fluxnet package's flux_read() at daily resolution.  We keep loading the
# DD file here so Chunk 6 requires no modification, and name it `usmms_raw` to
# preserve the variable name that Chunk 6 uses.
usmms_raw <- read_csv("../data/US-MMS/US-MMS_DD.csv", show_col_types = FALSE)

# ── CMIP6 monthly output (all three models) — unchanged ──────────────────────
# Pre-extracted from the Pangeo CMIP6 catalog at the grid cell nearest to
# US-MMS. One row per month; units for carbon fluxes: kg C m⁻² s⁻¹.
cesm2_raw   <- read_csv("../data/cmip6/CESM2_usmms_monthly.csv",
                        show_col_types = FALSE)
ipsl_raw    <- read_csv("../data/cmip6/IPSL-CM6A-LR_usmms_monthly.csv",
                        show_col_types = FALSE)
ukesm_raw   <- read_csv("../data/cmip6/UKESM1-0-LL_usmms_monthly.csv",
                        show_col_types = FALSE)

# ── Discovery: print the unit and format situation ────────────────────────────
cat("── Sub-daily FLUXNET: loaded file sizes ────────────────────────────────\n")
cat(sprintf("  DE-Tha HH rows: %s (1996–2024, 30 min)\n",
            format(nrow(detha_hh_raw), big.mark = ",")))
cat(sprintf("  DK-Sor HH rows: %s (1996–2024, 30 min)\n",
            format(nrow(dksor_hh_raw), big.mark = ",")))
cat(sprintf("  US-MMS HR rows: %s (1999–2023, 60 min)\n",
            format(nrow(usmms_hr_raw), big.mark = ",")))
cat("\n── Sub-daily FLUXNET: carbon flux columns ──────────────────────────────\n")
cat("  Variable          Raw units             Notes\n")
cat("  NEE_VUT_REF       μmol CO₂ m⁻² s⁻¹    Instantaneous rate; + = source\n")
cat("  NEE_VUT_REF_QC    0–3 integer          0=measured, 1–3=gap-fill tier\n")
cat("  Timestamp column:\n")
cat("    DE-Tha / DK-Sor:  DATETIME_START  (ISO 8601, e.g. 1996-01-01T00:00:00Z)\n")
cat("    US-MMS:           TIMESTAMP_START (integer YYYYMMDDHHMM, e.g. 199901010000)\n")
cat("  The timestamp difference is handled in Chunk 3 (harmonize).\n\n")
cat("── CMIP6 monthly: key columns ──────────────────────────────────────────\n")
cat("  gpp_kgC_m2_s     kg C m⁻² s⁻¹  (monthly average RATE — very different\n")
cat("                                   from FLUXNET μmol units; see Chunk 3)\n\n")
cat("── SSEM output (from Exercise 01) ──────────────────────────────────────\n")
cat("  GPP [dim 3 = 5]  μmol CO₂ m⁻² s⁻¹  (30-min average rate)\n")
cat("  NEP [dim 3 = 6]  μmol CO₂ m⁻² s⁻¹  (positive = ecosystem gaining C)\n")
cat("  Negate NEP → NEE so sign convention matches FLUXNET (done in Chunk 3).\n")
```

---

## Unit harmonization

> **A note on embedded assumptions.** Every unit conversion below is also a
> scientific decision. Multiplying a half-hourly rate by 1800 seconds assumes
> the rate was constant over that 30-minute window — true for FLUXNET gap-filled
> data by construction, but an approximation for the model. Choosing
> `GPP_NT_VUT_REF` (nighttime-partitioning method) rather than
> `GPP_DT_VUT_REF` (daytime method) can change the annual GPP estimate by
> 5–15% at individual sites. Negating SSEM's NEP to produce an NEE-convention
> value embeds the assumption that storage fluxes are negligible — reasonable
> at the daily timescale, questionable at sub-hourly timescales. Read each
> conversion below and make sure you understand both the arithmetic and the
> assumption it encodes.

```{r harmonize, message=FALSE}
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# CONVERSION CONSTANTS (centralised per CLAUDE.md — define once, use everywhere)
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

# Molar mass of carbon in grams per micromole CO₂
MC <- 12e-6   # gC μmol⁻¹

# Seconds per minute
SECS_PER_MIN <- 60L

# Per-site conversion: μmol CO₂ m⁻² s⁻¹ (instantaneous rate) → gC m⁻² per step.
# Formula: MC × timestep_length_in_minutes × SECS_PER_MIN
# = MC × timestep_in_seconds
#
# SCIENTIFIC ASSUMPTION EMBEDDED HERE:
#   This conversion assumes the flux rate is constant over the entire timestep.
#   For FLUXNET gap-filled data this is true by construction: each 30-min or
#   60-min value IS the mean over that window. For SSEM output it is also true
#   by design (each timestep produces a single mean rate). The assumption
#   becomes an approximation only when up-scaling from fine to coarse time steps.
#
# DE-Tha / DK-Sor (30-min): 12e-6 × 30 × 60 = 0.0216 gC per step
# US-MMS         (60-min): 12e-6 × 60 × 60 = 0.0432 gC per step
#
# Using the same constant for both would halve or double one site's daily totals.
# This would make two ecologically similar forests appear to have very different
# carbon balances for a purely methodological reason — a common error in
# multi-site model-data comparison papers.
umol_s_to_gC_step <- function(site) {
  MC * site_timestep[[site]] * SECS_PER_MIN
}

# Daily average rate → daily total (used for CMIP6 comparisons in Chunk 6)
UMOL_S_TO_GC_DAY <- MC * 86400            # μmol m⁻² s⁻¹ → gC m⁻² day⁻¹ = 1.0368
KGC_S_TO_GC_DAY  <- 1000 * 86400          # kg C m⁻² s⁻¹ → gC m⁻² day⁻¹ = 86400000
MGC_HA_TO_KGC_M2 <- 1000 / 10000          # Mg C ha⁻¹ → kg C m⁻² = 0.1

# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# PART A — SSEM: produce both sub-daily and daily representations
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# SSEM was run for DE-Tha year 2000 at 30-minute resolution (see Exercise 01).
# Year 2000 is a leap year: 366 days × 48 steps/day = 17,568 timesteps.
# We build two representations:
#   ssem_subdaily: one row per 30-min step (for sub-daily comparison in Chunk 4)
#   ssem_daily:    one row per calendar day (for annual cycle and cost functions)

n_steps        <- dim(ssem_output)[1]            # 17,568
ssem_step_mins <- site_timestep[["DE-Tha"]]      # 30 minutes

# Build a POSIXct datetime vector aligned to FLUXNET HH timestamps.
# FLUXNET DATETIME_START represents the start of each 30-min window.
# The first driver row loaded in Exercise 01 was 2000-01-01 00:00 UTC.
ssem_datetime <- seq(
  from       = as.POSIXct("2000-01-01 00:00:00", tz = "UTC"),
  by         = paste(ssem_step_mins, "mins"),
  length.out = n_steps
)

# Variable 6 in SSEM output is NEP (net ecosystem production, positive = C sink).
# Negate to produce NEE (net ecosystem exchange, positive = atmospheric source)
# so SSEM and FLUXNET use the same sign convention throughout this exercise.
ssem_nee_raw <- -ssem_output[,, 6]   # [17568 × 100], μmol CO₂ m⁻² s⁻¹

# Sub-daily summary: ensemble 2.5%, 50%, 97.5% at each 30-min step.
ssem_sd_ci <- apply(ssem_nee_raw, 1, quantile, c(0.025, 0.5, 0.975), na.rm = TRUE)

ssem_subdaily <- tibble(
  datetime = ssem_datetime,
  date     = as.Date(ssem_datetime),
  site     = "DE-Tha",
  nee_med  = ssem_sd_ci["50%",   ],   # μmol CO₂ m⁻² s⁻¹
  nee_lo   = ssem_sd_ci["2.5%",  ],
  nee_hi   = ssem_sd_ci["97.5%", ]
)

# Daily aggregation: sum per-step NEE contributions within each calendar day.
# Rate (μmol m⁻² s⁻¹) × step conversion (gC/step) summed over 48 steps/day
# gives daily total in gC m⁻² day⁻¹.
step_conv <- umol_s_to_gC_step("DE-Tha")   # 0.0216 gC per 30-min step
n_days    <- n_steps / 48
day_idx   <- rep(seq_len(n_days), each = 48)
ssem_dates <- seq.Date(as.Date("2000-01-01"), by = "day", length.out = n_days)

daily_nee_ens <- apply(ssem_nee_raw, 2, function(ens) {
  tapply(ens, day_idx, sum) * step_conv
})   # [366 × 100]

daily_gpp_ens <- apply(ssem_output[,, 5], 2, function(ens) {
  tapply(ens, day_idx, sum) * step_conv
})

nee_ci <- apply(daily_nee_ens, 1, quantile, c(0.025, 0.5, 0.975), na.rm = TRUE)
gpp_ci <- apply(daily_gpp_ens, 1, quantile, c(0.025, 0.5, 0.975), na.rm = TRUE)

ssem_daily <- tibble(
  date    = ssem_dates,
  site    = "DE-Tha",
  nee_med = nee_ci["50%",   ],   # gC m⁻² day⁻¹
  nee_lo  = nee_ci["2.5%",  ],
  nee_hi  = nee_ci["97.5%", ],
  gpp_med = gpp_ci["50%",   ],
  gpp_lo  = gpp_ci["2.5%",  ],
  gpp_hi  = gpp_ci["97.5%", ]
)

cat("SSEM sub-daily rows:", nrow(ssem_subdaily),
    "| SSEM daily rows:", nrow(ssem_daily),
    "| Annual GPP median (gC m⁻² yr⁻¹):",
    round(sum(ssem_daily$gpp_med), 0), "\n")

# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# PART B — FLUXNET: standardise sub-daily HH/HR data, then aggregate to daily
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

# Step 1: parse timestamps and select key columns.
# The timestamp column differs between sites:
#   DE-Tha / DK-Sor: DATETIME_START (ISO 8601 with trailing Z = UTC)
#   US-MMS:          TIMESTAMP_START (integer YYYYMMDDHHMM, raw FLUXNET format)
harmonise_subdaily <- function(df, site_label) {
  if ("DATETIME_START" %in% names(df)) {
    dt <- as.POSIXct(df$DATETIME_START, tz = "UTC")
  } else {
    # YYYYMMDDHHMM integer → POSIXct.
    # as.character() handles integer64 (bit64) safely.
    dt <- as.POSIXct(as.character(df$TIMESTAMP_START),
                     format = "%Y%m%d%H%M", tz = "UTC")
  }
  tibble(
    datetime = dt,
    date     = as.Date(dt),
    site     = site_label,
    nee_raw  = df$NEE_VUT_REF,       # μmol CO₂ m⁻² s⁻¹, source-positive
    nee_qc   = df$NEE_VUT_REF_QC     # 0 = measured, 1–3 = gap-fill tiers
  )
}

detha_subdaily <- harmonise_subdaily(detha_hh_raw, "DE-Tha")
dksor_subdaily <- harmonise_subdaily(dksor_hh_raw, "DK-Sor")
usmms_subdaily <- harmonise_subdaily(usmms_hr_raw, "US-MMS")

fluxnet_subdaily <- bind_rows(detha_subdaily, dksor_subdaily, usmms_subdaily)

# Step 2: aggregate to daily totals using the correct per-site step length.
# Only compute a daily total if at least 75% of expected timesteps are non-NA.
# Fewer than 75% means the sum would be artificially low and could make one day
# look like a weaker sink than it actually was.
aggregate_to_daily <- function(df, site_label) {
  n_expected <- 1440L / site_timestep[[site_label]]   # 48 or 24 steps per day
  step_gC    <- umol_s_to_gC_step(site_label)

  df |>
    filter(site == site_label) |>
    group_by(date, site) |>
    summarise(
      n_steps_ok  = sum(!is.na(nee_raw)),
      nee_gCm2d   = if_else(
        n_steps_ok >= 0.75 * n_expected,
        sum(nee_raw * step_gC, na.rm = TRUE),   # daily total in gC m⁻² day⁻¹
        NA_real_
      ),
      nee_qc_mean = mean(nee_qc, na.rm = TRUE),  # for cost-function filtering
      .groups     = "drop"
    )
}

fluxnet_daily <- bind_rows(
  aggregate_to_daily(fluxnet_subdaily, "DE-Tha"),
  aggregate_to_daily(fluxnet_subdaily, "DK-Sor"),
  aggregate_to_daily(fluxnet_subdaily, "US-MMS")
)

# Step 3: identify the most recent complete year for each site.
# "Complete" = at least 330 days in the year have a non-NA daily NEE total.
# This is used in Chunk 4 to choose which annual cycle to display for each site.
# DE-Tha: most recent year used for the sub-daily July panel is forced to 2000
# by the SSEM run; for other sites we use their own most recent complete year.
most_recent_complete_year <- function(site_label) {
  fluxnet_daily |>
    filter(site == site_label) |>
    mutate(yr = year(date)) |>
    group_by(yr) |>
    summarise(n_ok = sum(!is.na(nee_gCm2d)), .groups = "drop") |>
    filter(n_ok >= 330) |>
    pull(yr) |>
    max()
}

obs_year <- list(
  "DE-Tha" = 2000L,                                      # fixed: SSEM run year
  "DK-Sor" = most_recent_complete_year("DK-Sor"),
  "US-MMS" = most_recent_complete_year("US-MMS")
)

cat("Observation year used for July sub-daily comparison and annual cycle:\n")
for (s in names(obs_year)) cat(sprintf("  %s: %d\n", s, obs_year[[s]]))

# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# PART C — CMIP6: convert kg C m⁻² s⁻¹ → gC m⁻² day⁻¹ (unchanged)
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━

harmonise_cmip6 <- function(df, model_label) {
  df |>
    mutate(
      model      = model_label,
      gpp_gCm2d  = gpp_kgC_m2_s * KGC_S_TO_GC_DAY,
      rh_gCm2d   = rh_kgC_m2_s  * KGC_S_TO_GC_DAY,
      ra_gCm2d   = ra_kgC_m2_s  * KGC_S_TO_GC_DAY,
      days_in_mo = days_in_month(as.Date(paste(year, month, "01", sep = "-")))
    ) |>
    select(model, year, month, days_in_mo, scenario,
           gpp_gCm2d, rh_gCm2d, ra_gCm2d, lai_m2_m2)
}

cesm2     <- harmonise_cmip6(cesm2_raw,  "CESM2")
ipsl      <- harmonise_cmip6(ipsl_raw,   "IPSL-CM6A-LR")
ukesm     <- harmonise_cmip6(ukesm_raw,  "UKESM1-0-LL")
cmip6_all <- bind_rows(cesm2, ipsl, ukesm)

# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# PART D — FLUXCOM placeholder (unchanged from original)
# ━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━
# Representative annual GPP from Nelson et al. (2024) for the DE-Tha region.
# Replace with values read from data/fluxcom/ when the file is available.
fluxcom_gpp_annual <- 1080   # gC m⁻² yr⁻¹
fluxcom_gpp_sd     <- 130    # ±1 SD from FLUXCOM ensemble

cat("\nHarmonization complete.\n")
cat(sprintf("  FLUXNET sub-daily rows: %s\n", format(nrow(fluxnet_subdaily), big.mark=",")))
cat(sprintf("  FLUXNET daily rows:     %s\n", format(nrow(fluxnet_daily),    big.mark=",")))
cat(sprintf("  CMIP6 monthly rows:     %d\n", nrow(cmip6_all)))
```

---

## Visual comparison

```{r compare_subdaily, fig.width=11, fig.height=8}
#| fig.cap: >
#|   Sub-daily NEE during July at each site. Blue ribbon: SSEM 95% credible
#|   interval (DE-Tha only). Blue line: SSEM ensemble median. Grey points:
#|   FLUXNET NEE_VUT_REF at native sub-daily resolution. SSEM was run only at
#|   DE-Tha for year 2000; DK-Sor and US-MMS show observations from their most
#|   recent complete year (see strip labels). Units: μmol CO₂ m⁻² s⁻¹; positive
#|   = atmospheric source. Note that the US-MMS panel has half as many points
#|   per day (hourly vs half-hourly) as the European sites.

# ── Panel 1: sub-daily NEE for July ──────────────────────────────────────────
# July is the peak-growing-season month at all three sites, making the diurnal
# cycle most pronounced and differences in GPP magnitude most visible.
# Negative values (mid-day sink) dominate during daytime; respiration drives
# positive values at night. SSEM's light-use-efficiency model captures the
# daytime dip but cannot represent water-stress-driven afternoon depression
# in GPP — watch for that gap between SSEM and observations on hot days.

# SSEM: extract July 2000 sub-daily steps
ssem_july <- ssem_subdaily |>
  filter(month(datetime) == 7L)   # month() returns integer; year is already 2000

# FLUXNET: each site uses its own obs_year, filtered to July
fluxnet_july <- fluxnet_subdaily |>
  mutate(yr = year(datetime)) |>
  filter(
    (site == "DE-Tha" & yr == obs_year[["DE-Tha"]]) |
    (site == "DK-Sor" & yr == obs_year[["DK-Sor"]]) |
    (site == "US-MMS" & yr == obs_year[["US-MMS"]])
  ) |>
  filter(month(datetime) == 7L)

# Facet strip labels: embed the year and vegetation type so students know
# which year each panel represents without reading the caption.
july_labels <- c(
  "DE-Tha" = sprintf("DE-Tha — Norway spruce (DE) — July %d + SSEM",
                     obs_year[["DE-Tha"]]),
  "DK-Sor" = sprintf("DK-Sor — European beech (DK) — July %d",
                     obs_year[["DK-Sor"]]),
  "US-MMS" = sprintf("US-MMS — mixed deciduous (IN) — July %d  [hourly]",
                     obs_year[["US-MMS"]])
)

ggplot(mapping = aes(x = datetime)) +
  # SSEM 95% credible interval — steelblue ribbon (DE-Tha only)
  geom_ribbon(
    data = ssem_july,
    aes(ymin = nee_lo, ymax = nee_hi),
    fill = "steelblue", alpha = 0.30
  ) +
  # SSEM ensemble median line
  geom_line(
    data = ssem_july,
    aes(y = nee_med),
    colour = "steelblue3", linewidth = 0.5
  ) +
  # FLUXNET observed sub-daily NEE (all sites)
  geom_point(
    data    = fluxnet_july,
    mapping = aes(y = nee_raw),
    colour  = "grey20", alpha = 0.30, size = 0.3
  ) +
  # Zero reference: above = source, below = sink
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50",
             linewidth = 0.35) +
  facet_wrap(
    ~ factor(site, levels = names(july_labels), labels = july_labels),
    ncol = 1, scales = "free_x"
  ) +
  labs(
    title    = "Sub-daily NEE — July comparison across three temperate forest sites",
    subtitle = paste("Units: μmol CO₂ m⁻² s⁻¹  |  + = source  |",
                     "SSEM shown for DE-Tha 2000 only"),
    x        = NULL,
    y        = expression(NEE ~ (μmol ~ CO[2] ~ m^{-2} ~ s^{-1}))
  ) +
  theme_classic(base_size = 12) +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold", size = 9),
        panel.spacing     = unit(1.2, "lines"))
```

```{r compare_daily_annual, fig.width=11, fig.height=8}
#| fig.cap: >
#|   Annual cycle of daily NEE for each site. Blue ribbon: SSEM 95% credible
#|   interval (DE-Tha 2000 only). Blue line: SSEM ensemble median. Grey points:
#|   FLUXNET daily NEE_VUT_REF aggregated from sub-daily data. Each site shows
#|   its most recent complete year (see strip labels). x-axis is day of year so
#|   all sites share the same seasonal scale regardless of calendar year.
#|   Units: gC m⁻² day⁻¹; positive = source.

# ── Panel 2: full annual cycle at daily aggregated scale ──────────────────────
# The sub-daily July panel shows the fine-scale dynamics; this panel puts them
# in annual context. Students should notice: (a) how much of the annual carbon
# budget is concentrated in the summer months, (b) whether SSEM's spring
# transition timing matches the observations at DE-Tha, and (c) the contrast
# between evergreen (DE-Tha, flatter seasonal cycle) and deciduous
# (DK-Sor, US-MMS, sharper spring green-up) vegetation types.

# Filter each site to its designated observation year
annual_fluxnet <- fluxnet_daily |>
  mutate(yr = year(date)) |>
  filter(
    (site == "DE-Tha" & yr == obs_year[["DE-Tha"]]) |
    (site == "DK-Sor" & yr == obs_year[["DK-Sor"]]) |
    (site == "US-MMS" & yr == obs_year[["US-MMS"]])
  ) |>
  mutate(doy = yday(date))

# Add day-of-year to SSEM for the shared x-axis
ssem_doy <- ssem_daily |>
  mutate(doy = yday(date))

# Strip labels with year context
annual_labels <- c(
  "DE-Tha" = sprintf("DE-Tha — Norway spruce (DE) — %d", obs_year[["DE-Tha"]]),
  "DK-Sor" = sprintf("DK-Sor — European beech (DK) — %d", obs_year[["DK-Sor"]]),
  "US-MMS" = sprintf("US-MMS — mixed deciduous (IN) — %d", obs_year[["US-MMS"]])
)

ggplot(mapping = aes(x = doy)) +
  # SSEM uncertainty ribbon and median (DE-Tha only)
  geom_ribbon(
    data = ssem_doy,
    aes(ymin = nee_lo, ymax = nee_hi),
    fill = "steelblue", alpha = 0.30
  ) +
  geom_line(
    data = ssem_doy,
    aes(y = nee_med),
    colour = "steelblue3", linewidth = 0.5
  ) +
  # FLUXNET daily NEE
  geom_point(
    data    = annual_fluxnet,
    mapping = aes(y = nee_gCm2d),
    colour  = "grey20", alpha = 0.40, size = 0.6
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50",
             linewidth = 0.35) +
  facet_wrap(
    ~ factor(site, levels = names(annual_labels), labels = annual_labels),
    ncol = 1, scales = "free_y"
  ) +
  scale_x_continuous(
    breaks = c(1, 60, 121, 182, 244, 305, 366),
    labels = c("Jan", "Mar", "May", "Jul", "Sep", "Nov", "Dec")
  ) +
  labs(
    title    = "Annual cycle of daily NEE — most recent complete year per site",
    subtitle = paste("Units: gC m⁻² day⁻¹  |  + = source  |",
                     "SSEM shown for DE-Tha 2000 only"),
    x        = "Day of year",
    y        = expression(NEE ~ (gC ~ m^{-2} ~ day^{-1}))
  ) +
  theme_classic(base_size = 12) +
  theme(strip.background = element_blank(),
        strip.text        = element_text(face = "bold", size = 9),
        panel.spacing     = unit(1.2, "lines"))
```

---

## Cost functions

```{r cost_functions}
# ── Cost function metrics at sub-daily and daily scales ───────────────────────
# Three metrics capture different aspects of model performance:
#
#   RMSE  — root mean square error: the typical magnitude of a single error.
#           Penalises large errors quadratically, so it is dominated by the
#           worst predictions. Units match the variable (μmol m⁻² s⁻¹ at
#           sub-daily scale; gC m⁻² day⁻¹ at daily scale — not comparable
#           in magnitude, but dimensionless r is directly comparable).
#
#   Bias  — mean(model − observation): a non-zero bias indicates a systematic
#           offset that RMSE cannot distinguish from random scatter. Positive
#           bias means the model predicts a larger source (or smaller sink)
#           on average. Often caused by a mismatch in either GPP or Reco.
#
#   r     — Pearson correlation coefficient: whether the model tracks the
#           temporal pattern of observations, independent of absolute level.
#           A model can achieve r = 0.9 while having a large RMSE if the
#           seasonal shape is right but the amplitude is wrong.
#
# TIMESCALE DEPENDENCY — a modeling choice:
#   Cost function values change with the timescale at which they are computed.
#   Daily metrics smooth over the diurnal cycle, reducing sensitivity to the
#   representation of photosynthesis onset, stomatal responses, and nighttime
#   respiration. Sub-daily metrics expose these processes directly.
#   Neither timescale is universally "correct" — the right choice depends on
#   the management or policy question the model is meant to answer:
#     • Sub-daily: relevant if the model is used for flux partitioning,
#       phenology detection, or land–atmosphere coupling in an ESM.
#     • Daily: relevant if the model is used for seasonal carbon budgets,
#       forest management planning, or carbon credit accounting.
#   Reporting metrics at both scales is good practice because it prevents a
#   model from appearing skillful at one scale while hiding failures at another.

# ── Helper: compute RMSE, bias, and Pearson r ─────────────────────────────────
compute_metrics <- function(model_vec, obs_vec) {
  valid <- !is.na(model_vec) & !is.na(obs_vec)
  n     <- sum(valid)
  if (n < 2L) return(list(n = 0L, rmse = NA_real_, bias = NA_real_, r = NA_real_))
  res <- model_vec[valid] - obs_vec[valid]
  list(
    n    = n,
    rmse = sqrt(mean(res^2)),
    bias = mean(res),
    r    = cor(model_vec[valid], obs_vec[valid])
  )
}

# ── Sub-daily metrics (μmol CO₂ m⁻² s⁻¹) ────────────────────────────────────
# Join SSEM sub-daily timesteps with FLUXNET HH observations for DE-Tha 2000.
# Filter to well-measured timesteps: QC flag 0 (measured) or 1 (good gap-fill).
# QC 2 and 3 indicate coarser gap-filling where the "observation" is largely
# model-derived — including those would inflate r artificially.
detha_sd_eval <- detha_subdaily |>
  filter(year(datetime) == 2000L, nee_qc <= 1L) |>
  inner_join(
    ssem_subdaily |> select(datetime, nee_med),
    by = "datetime"
  )

sd_de  <- compute_metrics(detha_sd_eval$nee_med, detha_sd_eval$nee_raw)

# DK-Sor and US-MMS: SSEM was not run at these sites, so no model values exist.
# We include NA rows in the table to make this explicit rather than omitting
# those sites — the absence of SSEM output is itself a result that tells
# students the model has not been evaluated there.
sd_dk  <- list(n = NA_integer_, rmse = NA_real_, bias = NA_real_, r = NA_real_)
sd_us  <- list(n = NA_integer_, rmse = NA_real_, bias = NA_real_, r = NA_real_)

# ── Daily metrics (gC m⁻² day⁻¹) ─────────────────────────────────────────────
# Filter to days where the mean HH QC flag <= 1 (most timesteps were measured).
# This is stricter than filtering at QC >= 0.5 (the CLAUDE.md standard for
# daily data), and reflects that we are now aggregating from sub-daily QC flags.
detha_dd_eval <- fluxnet_daily |>
  filter(site == "DE-Tha", year(date) == 2000L,
         !is.na(nee_gCm2d), nee_qc_mean <= 1.0) |>
  inner_join(ssem_daily |> select(date, nee_med), by = "date")

dd_de  <- compute_metrics(detha_dd_eval$nee_med, detha_dd_eval$nee_gCm2d)
dd_dk  <- list(n = NA_integer_, rmse = NA_real_, bias = NA_real_, r = NA_real_)
dd_us  <- list(n = NA_integer_, rmse = NA_real_, bias = NA_real_, r = NA_real_)

# ── Summary table ─────────────────────────────────────────────────────────────
metrics_tbl <- tibble(
  Site        = rep(c("DE-Tha", "DK-Sor", "US-MMS"), each = 2L),
  Timescale   = rep(c("Sub-daily", "Daily"), times = 3L),
  Units       = rep(c("μmol m⁻² s⁻¹", "gC m⁻² day⁻¹"), times = 3L),
  N           = c(sd_de$n,  dd_de$n,  sd_dk$n,  dd_dk$n,  sd_us$n,  dd_us$n),
  RMSE        = c(sd_de$rmse, dd_de$rmse, sd_dk$rmse, dd_dk$rmse,
                  sd_us$rmse, dd_us$rmse),
  Bias        = c(sd_de$bias, dd_de$bias, sd_dk$bias, dd_dk$bias,
                  sd_us$bias, dd_us$bias),
  Correlation = c(sd_de$r,  dd_de$r,  sd_dk$r,  dd_dk$r,  sd_us$r,  dd_us$r)
)

cat("── Cost function metrics: SSEM ensemble median vs FLUXNET NEE ──────────\n")
cat("   SSEM was run at DE-Tha year 2000 only.\n")
cat("   NA = no SSEM output available for this site × timescale combination.\n")
cat("   Sub-daily units: μmol CO₂ m⁻² s⁻¹\n")
cat("   Daily units:     gC m⁻² day⁻¹  (not directly comparable in magnitude)\n\n")

metrics_tbl |>
  mutate(across(c(RMSE, Bias, Correlation), \(x) round(x, 3))) |>
  print(n = Inf)

cat("\nParticle filter connection (afternoon session):\n")
cat("  Under a Gaussian error model with constant observation uncertainty σ,\n")
cat("  maximising log-likelihood is equivalent to minimising the sum of\n")
cat("  squared residuals — the quantity RMSE² × N summarises. The particle\n")
cat("  filter goes further: it weights each ensemble member by its full\n")
cat("  time-series likelihood, folds in the prior (parameter distributions),\n")
cat("  and propagates the posterior forward. The timescale at which you\n")
cat("  evaluate the likelihood is therefore part of the PF design.\n")

# ── Seasonal residual plot at daily scale ─────────────────────────────────────
# Plotting residuals by day of year reveals whether errors are systematic
# across seasons or concentrated in a particular period. A smooth LOESS trend
# that departs from zero in summer suggests a missing warm-season process
# (water stress, VPD limitation, phenology); a winter departure implicates
# the temperature-respiration relationship.
detha_dd_eval |>
  mutate(residual = nee_med - nee_gCm2d,
         doy      = yday(date)) |>
  ggplot(aes(x = doy, y = residual)) +
    geom_point(alpha = 0.40, size = 0.8, colour = "tomato3") +
    geom_smooth(method = "loess", span = 0.3, se = FALSE,
                colour = "black", linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      title    = "Seasonal residuals: SSEM − FLUXNET daily NEE at DE-Tha 2000",
      subtitle = "Positive: model predicts larger source than observed | QC mean ≤ 1 days only",
      x        = "Day of year",
      y        = expression(Residual ~ (gC ~ m^{-2} ~ day^{-1}))
    ) +
    theme_classic(base_size = 12)
```

---

## CMIP6 comparison

> **Choose your model.** Set `chosen_model` to one of:
> `"CESM2"`, `"IPSL-CM6A-LR"`, `"UKESM1-0-LL"`
>
> Each represents a different Earth system model from the CMIP6 multi-model
> ensemble. CESM2 is NCAR's community model; IPSL-CM6A-LR is the Institut
> Pierre-Simon Laplace model; UKESM1-0-LL is the UK Earth System Model.
> All three are calibrated against global carbon budgets and observations,
> not against individual flux tower records.

```{r cmip6_compare, fig.width=10, fig.height=6}
#| fig.cap: >
#|   Monthly GPP and heterotrophic respiration (Rh) from the selected CMIP6
#|   model (coloured lines) versus FLUXNET monthly means at US-MMS (black
#|   points). FLUXNET RECO (total ecosystem respiration = Ra + Rh) is shown
#|   rather than Rh alone because FLUXNET does not partition autotrophic and
#|   heterotrophic respiration separately; expect RECO > CMIP6 Rh. Both axes
#|   show gC m⁻² day⁻¹ (monthly average rate), covering 1999–2014.

# ── USER SELECTION: change this line to choose a different model ──────────────
chosen_model <- "CESM2"   # options: "CESM2", "IPSL-CM6A-LR", "UKESM1-0-LL"
# ─────────────────────────────────────────────────────────────────────────────

# Validate the choice so students get a clear error message if they mistype
stopifnot(
  "`chosen_model` must be 'CESM2', 'IPSL-CM6A-LR', or 'UKESM1-0-LL'" =
    chosen_model %in% c("CESM2", "IPSL-CM6A-LR", "UKESM1-0-LL")
)
cat("Comparing against:", chosen_model, "\n")

# Extract the chosen model and add a proper date column (first day of each month)
cmip6_chosen <- cmip6_all |>
  filter(model == chosen_model) |>
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))

# ── Aggregate FLUXNET US-MMS to monthly means (gC m⁻² day⁻¹) ─────────────────
# FLUXNET US-MMS data starts 1999 and the CMIP6 historical run covers 1980–2014.
# We use the full overlap: 1999–2014.
# Taking the monthly MEAN of daily gC m⁻² day⁻¹ values gives a rate on the
# same time axis as CMIP6 (which also reports monthly average daily rates).
usmms_monthly <- usmms_raw |>
  mutate(date = as.Date(DATE)) |>
  filter(year(date) >= 1999, year(date) <= 2014) |>
  mutate(across(where(is.numeric), \(x) replace(x, x == -9999, NA))) |>
  mutate(
    gpp_gCm2d  = GPP_NT_VUT_REF  * UMOL_S_TO_GC_DAY,
    reco_gCm2d = RECO_NT_VUT_REF * UMOL_S_TO_GC_DAY
  ) |>
  group_by(year = year(date), month = month(date)) |>
  summarise(
    gpp_gCm2d  = mean(gpp_gCm2d,  na.rm = TRUE),
    reco_gCm2d = mean(reco_gCm2d, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  mutate(date = as.Date(paste(year, month, "01", sep = "-")))

# ── Plot monthly GPP: CMIP6 vs FLUXNET ───────────────────────────────────────
# Pivot both sources to long format so ggplot can colour them separately.
cmip6_long <- cmip6_chosen |>
  select(date, GPP = gpp_gCm2d, Rh = rh_gCm2d) |>
  pivot_longer(c(GPP, Rh), names_to = "variable", values_to = "value") |>
  mutate(source = chosen_model)

flux_long <- usmms_monthly |>
  # Rename RECO to "Rh" for the plot panel — but annotate the mismatch
  select(date, GPP = gpp_gCm2d, Rh = reco_gCm2d) |>
  pivot_longer(c(GPP, Rh), names_to = "variable", values_to = "value") |>
  mutate(source = "FLUXNET US-MMS")

bind_rows(cmip6_long, flux_long) |>
  ggplot(aes(x = date, y = value, colour = source)) +
    geom_line(linewidth = 0.4, alpha = 0.7) +
    # Annual mean smoothed to show interannual trend
    geom_smooth(method = "loess", span = 0.15, se = FALSE,
                linewidth = 1.0, na.rm = TRUE) +
    facet_wrap(~ variable, ncol = 1, scales = "free_y",
               labeller = as_labeller(c(
                 GPP = "GPP (gC m⁻² day⁻¹)",
                 Rh  = "Rh / RECO (gC m⁻² day⁻¹) — see note on scale mismatch"
               ))) +
    scale_colour_manual(
      values = c("FLUXNET US-MMS" = "grey20",
                 "CESM2"          = "#E07B39",
                 "IPSL-CM6A-LR"   = "#5B8DB8",
                 "UKESM1-0-LL"    = "#6BAE75")[c("FLUXNET US-MMS", chosen_model)]
    ) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(
      title    = paste("Monthly GPP and Rh —", chosen_model, "vs FLUXNET US-MMS"),
      subtitle = "1999–2014 | gC m⁻² day⁻¹ | Thick lines: LOESS trend",
      x        = NULL,
      y        = NULL,
      colour   = NULL
    ) +
    theme_classic(base_size = 12) +
    theme(
      legend.position  = "bottom",
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold")
    )
```

> **Scale mismatch note.** Each CMIP6 grid cell represents roughly 1° × 1° of
> latitude and longitude — approximately 100 km × 70 km at the latitude of
> US-MMS. That single number blends forest, cropland, urban surface, and open
> water, weighted by their fractional coverage in the land surface model.
> The flux tower footprint covers a few hundred metres to perhaps 1–2 km at
> most. The CMIP6 model was calibrated to reproduce global carbon budgets and
> continental-scale patterns, not to match individual tower records. Expecting
> it to agree with a tower to within 10% would be asking the wrong question.
> A better use of ESM output is to ask: does the model reproduce the *direction*
> of the seasonal cycle? Does its interannual variability correlate with
> observations? Does its decadal trend match the atmospheric record? FLUXCOM
> upscaled products, which blend tower observations with remote sensing and
> meteorology to produce gridded estimates at 0.5° resolution, are a better
> mediator for site-to-ESM comparisons because they operate at an intermediate
> spatial scale.
>
> **Further exploration.** Additional data for these sites are in:
>
> - `data/DK-Sor/` — DK-Sor daily, monthly, and yearly FLUXNET products
> - `data/US-MMS/` — US-MMS daily, monthly, and yearly products (no HH file)
>
> Try changing `chosen_model` above and re-knitting to compare models against
> each other. Try computing the same cost function metrics (Chunk 5) for the
> CMIP6 vs FLUXNET comparison and compare the numbers to SSEM.

---

## Discussion questions

Work through all four questions before you leave the session. These do not have
unique correct answers. The goal is to reason from the evidence in your plots
and tables.

**1. Which site does SSEM perform best on, and why?**
Look at the visual comparison and the cost function table. Does SSEM reproduce
the seasonal timing and magnitude of NEE at DE-Tha better or worse than you
expected, given that the parameters were chosen for that site? If you re-ran
SSEM with the same parameters at DK-Sor (a beech forest with strong spring
green-up) what specific failure mode would you predict, and which parameter
is most directly responsible?

**2. What does the seasonal pattern of residuals tell you about what is
missing?**
Look at the residual plot in Chunk 5. Are the errors random throughout the
year, or do they have a systematic seasonal structure? If they are larger in
summer than in winter, or positive in spring and negative in autumn, what
process does that implicate? Consider the list of things SSEM ignores — water
stress, VPD limitation, phenology — and identify which omission is most
consistent with the pattern you observe.

**3. Would you trust a forecast from this model?**
Imagine a stakeholder (a forest manager, a carbon credit verification service,
a national carbon inventory) asked you to use SSEM to project annual NEE five
years into the future for DE-Tha under a warming scenario. Based on the
validation you have done, what would you tell them? Be specific: what does the
model get right that you could defend, what does it get wrong that you could not
defend, and what additional validation steps would you require before presenting
the forecast as credible?

**4. What would you need to change about SSEM to improve performance at a
deciduous broadleaf forest (DBF) site?**
US-MMS and DK-Sor are both deciduous — they have near-zero GPP in winter and
a rapid spring onset driven by photoperiod and temperature. SSEM has no
phenology module: Bleaf is driven entirely by the balance between NPPleaf
allocation and the constant litterfall rate, which produces a very slow,
temperature-driven seasonal cycle rather than a discrete green-up event. Sketch
(in words, not code) what equations you would need to add to SSEM to reproduce
a realistic deciduous phenology. What driver variables would those equations
require, and are those drivers available in the current data files?
