---
title: "Exercise 03 — AM/PM Handoff: Preparing the Calibration Workspace"
subtitle: "Fluxcourse 2026 · Bloomington, Indiana"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
---

```{r metadata, echo=TRUE}
# Reproducibility header — records when this document was knitted, the git
# commit it came from, and the full R session environment.
cat("Date:", format(Sys.time(), "%Y-%m-%d %H:%M"), "\n")
cat("Commit:", system("git rev-parse --short HEAD", intern = TRUE), "\n")
sessionInfo()
```

---

## Introduction

You have just finished the morning session. You ran the Super Simple Ecosystem
Model (SSEM) at two sites — US-NR1 (Niwot Ridge, an ENF) and US-MMS (Morgan
Monroe, a DBF) — and compared the output against observations, FLUXCOM, and
CMIP6 data. The afternoon session with Mike Dietze will take the next step:
using the particle filter to systematically adjust SSEM's parameters so they are
consistent with what the tower observed. That process is called **data
assimilation** or **calibration**.

This document bridges the two sessions. It prepares workspace files for one or
both primary sites — `exercises/output/handoff_workspace_NR1.RDS` and
`exercises/output/handoff_workspace_MMS.RDS` — that Mike's afternoon code will
load directly. Each workspace contains:

- The meteorological drivers for the selected site and year in the exact format
  SSEM expects (`inputs`)
- The prior parameter ensemble and initial conditions (`params`, `X.orig`)
- The prior SSEM ensemble forecast from those parameters (`output.ensemble`)
- The eddy covariance observations the particle filter will assimilate (`flux`,
  `nep`, `nep.qc`, `qaqc`)

**US-NR1** (subalpine ENF, Colorado) — half-hourly observations, 2008.
**US-MMS** (temperate DBF, Indiana) — hourly observations, the site you
visited on Friday; 2005 for the particle filter exercise (the year the PF
framework was originally developed and tested at this site).

**Your only task right now is to run this document.** There are no questions to
answer and no analysis to interpret. When the checklist at the bottom prints
without errors, raise your hand and let the instructor know. The afternoon
session will start from there.

---

## Setup

```{r setup, message=FALSE, warning=FALSE}
# ── Libraries ──────────────────────────────────────────────────────────────────
library(tidyverse)   # dplyr, ggplot2, readr, tidyr, lubridate wrapper
library(lubridate)   # POSIXct parsing, year(), hour()

# ── Paths ──────────────────────────────────────────────────────────────────────
# If your data are on Google Drive, replace data_dir below:
#   data_dir <- "/Volumes/GoogleDrive/My Drive/FluxCourse2026/data"
data_dir <- "../data"  # default for codespace
code_dir  <- "../R"

# ── Model code ─────────────────────────────────────────────────────────────────
source(file.path(code_dir, "functions.R"))
source(file.path(code_dir, "utils.R"))

# ── Handoff site selection ─────────────────────────────────────────────────────
# Set this to "US-NR1" or "US-MMS" depending on which site you want to hand
# off to the PM session.  This document will build and save the workspace for
# the selected site.  It will then also save the US-MMS workspace (the
# historically-used PM session site) so Mike can choose to use either one.
#
# "US-NR1" — subalpine ENF, Niwot Ridge, CO; half-hourly; year 2008
# "US-MMS" — temperate DBF, Morgan Monroe, IN; hourly; year 2005
handoff_site <- "US-NR1"   # change to "US-MMS" if you want the Indiana site
```

---

## Load observations — US-NR1 (ENF, year 2008)

```{r load_obs_nr1, message=FALSE}
# ── Path to the US-NR1 half-hourly data file ──────────────────────────────────
hr_file_nr1 <- file.path(data_dir, "US-NR1", "US-NR1_HR.csv")

if (!file.exists(hr_file_nr1)) {
  stop(
    "US-NR1 half-hourly data not found at: ", hr_file_nr1, "\n\n",
    "Run the following script from the repository root to download the data:\n",
    "  Rscript data/US-NR1/download_usnr1.R\n"
  )
}

message("Loading: ", basename(hr_file_nr1))
raw_nr1_full <- read_csv(
  hr_file_nr1,
  show_col_types = FALSE,
  col_select = any_of(c(
    "DATETIME_START",
    "SW_IN_F", "TA_F",
    "NEE_CUT_REF", "NEE_CUT_REF_QC",
    "NEE_CUT_REF_JOINTUNC"
  ))
)
cat(sprintf("Loaded %s rows × %d columns\n",
            format(nrow(raw_nr1_full), big.mark = ","), ncol(raw_nr1_full)))

# Check for joint uncertainty column
has_jointunc_nr1 <- "NEE_CUT_REF_JOINTUNC" %in% names(raw_nr1_full)
if (has_jointunc_nr1) {
  cat("NEE_CUT_REF_JOINTUNC found — will be included in workspace.\n")
} else {
  cat("NEE_CUT_REF_JOINTUNC not present. Creating NA placeholder column.\n")
}

# Replace FLUXNET missing-value sentinel with NA
raw_nr1_full <- raw_nr1_full |>
  mutate(across(where(is.numeric), \(x) replace(x, x == -9999, NA)))

# Parse ISO 8601 timestamp (same format as DE-Tha)
raw_nr1_full <- raw_nr1_full |>
  mutate(datetime = as.POSIXct(DATETIME_START, tz = "UTC")) |>
  filter(!is.na(datetime))

# Filter to year 2008
run_year_nr1 <- 2008L
raw_nr1 <- raw_nr1_full |> filter(year(datetime) == run_year_nr1)
cat(sprintf("Rows in %d: %s (expect 17,568 for a leap year at 30 min)\n",
            run_year_nr1, format(nrow(raw_nr1), big.mark = ",")))

# PAR conversion and gap-fill
raw_nr1 <- raw_nr1 |>
  mutate(
    PAR  = pmax(SW_IN_F * 2.1, 0),
    temp = TA_F
  ) |>
  mutate(PAR  = replace_na(PAR,  0)) |>
  tidyr::fill(temp, .direction = "downup")

# Trim to complete days
steps_per_day_nr1 <- 48L   # half-hourly
n_complete_nr1    <- (nrow(raw_nr1) %/% steps_per_day_nr1) * steps_per_day_nr1
if (n_complete_nr1 < nrow(raw_nr1)) {
  message(sprintf("Trimming %d partial rows.", nrow(raw_nr1) - n_complete_nr1))
  raw_nr1 <- raw_nr1[seq_len(n_complete_nr1), ]
}
n_days_nr1 <- nrow(raw_nr1) %/% steps_per_day_nr1
cat(sprintf("Complete days: %d  |  Timesteps: %d  |  Steps/day: %d\n",
            n_days_nr1, nrow(raw_nr1), steps_per_day_nr1))

# Build flux and inputs data frames for US-NR1
# Note: US-NR1 uses NEE_CUT_REF (constant u* threshold) instead of
# NEE_VUT_REF (variable u* threshold); see Exercise 02 for explanation.
flux_nr1 <- tibble(
  datetime             = raw_nr1$datetime,
  NEE_VUT_REF          = raw_nr1$NEE_CUT_REF,   # renamed for PF compatibility
  NEE_VUT_REF_QC       = raw_nr1$NEE_CUT_REF_QC,
  SW_IN_F              = raw_nr1$SW_IN_F,
  TA_F                 = raw_nr1$TA_F,
  NEE_VUT_REF_JOINTUNC = if (has_jointunc_nr1) raw_nr1$NEE_CUT_REF_JOINTUNC
                         else                   rep(NA_real_, nrow(raw_nr1))
)
inputs_nr1 <- tibble(
  date = raw_nr1$datetime,
  temp = raw_nr1$temp,
  PAR  = raw_nr1$PAR
)

cat(sprintf("flux_nr1: %d rows × %d columns\n", nrow(flux_nr1), ncol(flux_nr1)))
cat(sprintf("inputs_nr1: %d rows × %d columns\n", nrow(inputs_nr1), ncol(inputs_nr1)))
```

---

## Load observations — US-MMS (DBF, year 2005)

```{r load_obs_mms, message=FALSE}
# ── Path to the US-MMS hourly data file ───────────────────────────────────────
hr_file_mms <- file.path(data_dir, "US-MMS", "US-MMS_HR.csv")

if (!file.exists(hr_file_mms)) {
  stop(
    "US-MMS hourly data not found at: ", hr_file_mms, "\n\n",
    "Run the following scripts from the repository root:\n",
    "  Rscript data/US-MMS/download_usmms.R\n",
    "  Rscript data/US-MMS/extract_hr_workaround.R\n"
  )
}

message("Loading: ", basename(hr_file_mms))
raw_mms_full <- read_csv(
  hr_file_mms,
  show_col_types = FALSE,
  col_select = any_of(c(
    "TIMESTAMP_START",
    "SW_IN_F", "TA_F",
    "NEE_VUT_REF", "NEE_VUT_REF_QC",
    "NEE_VUT_REF_JOINTUNC"
  ))
)
cat(sprintf("Loaded %s rows × %d columns\n",
            format(nrow(raw_mms_full), big.mark = ","), ncol(raw_mms_full)))

has_jointunc_mms <- "NEE_VUT_REF_JOINTUNC" %in% names(raw_mms_full)
if (has_jointunc_mms) {
  cat("NEE_VUT_REF_JOINTUNC found — will be included in workspace.\n")
} else {
  cat("NEE_VUT_REF_JOINTUNC not present. Creating NA placeholder column.\n")
}

raw_mms_full <- raw_mms_full |>
  mutate(across(where(is.numeric), \(x) replace(x, x == -9999, NA)))

# US-MMS uses integer YYYYMMDDHHMM timestamp
raw_mms_full <- raw_mms_full |>
  mutate(
    datetime = as.POSIXct(as.character(TIMESTAMP_START),
                          format = "%Y%m%d%H%M", tz = "UTC")
  ) |>
  filter(!is.na(datetime))

# Filter to year 2005 (non-leap year, original PF development year)
run_year_mms <- 2005L
raw_mms <- raw_mms_full |> filter(year(datetime) == run_year_mms)
cat(sprintf("Rows in %d: %s (expect 8,760 for a non-leap year at hourly resolution)\n",
            run_year_mms, format(nrow(raw_mms), big.mark = ",")))

raw_mms <- raw_mms |>
  mutate(
    PAR  = pmax(SW_IN_F * 2.1, 0),
    temp = TA_F
  ) |>
  mutate(PAR  = replace_na(PAR,  0)) |>
  tidyr::fill(temp, .direction = "downup")

steps_per_day_mms <- 24L
n_complete_mms    <- (nrow(raw_mms) %/% steps_per_day_mms) * steps_per_day_mms
if (n_complete_mms < nrow(raw_mms)) {
  message(sprintf("Trimming %d partial rows.", nrow(raw_mms) - n_complete_mms))
  raw_mms <- raw_mms[seq_len(n_complete_mms), ]
}
n_days_mms <- nrow(raw_mms) %/% steps_per_day_mms
cat(sprintf("Complete days: %d  |  Timesteps: %d  |  Steps/day: %d\n",
            n_days_mms, nrow(raw_mms), steps_per_day_mms))

flux_mms <- tibble(
  datetime             = raw_mms$datetime,
  NEE_VUT_REF          = raw_mms$NEE_VUT_REF,
  NEE_VUT_REF_QC       = raw_mms$NEE_VUT_REF_QC,
  SW_IN_F              = raw_mms$SW_IN_F,
  TA_F                 = raw_mms$TA_F,
  NEE_VUT_REF_JOINTUNC = if (has_jointunc_mms) raw_mms$NEE_VUT_REF_JOINTUNC
                         else                   rep(NA_real_, nrow(raw_mms))
)
inputs_mms <- tibble(
  date = raw_mms$datetime,
  temp = raw_mms$temp,
  PAR  = raw_mms$PAR
)

cat(sprintf("flux_mms: %d rows × %d columns\n", nrow(flux_mms), ncol(flux_mms)))
cat(sprintf("inputs_mms: %d rows × %d columns\n", nrow(inputs_mms), ncol(inputs_mms)))
```

---

## Prior ensemble setup

```{r model_setup}
# ── Ensemble size ──────────────────────────────────────────────────────────────
ne <- 100L

# ── Timesteps (seconds per step) ──────────────────────────────────────────────
timestep_nr1 <- 1800L   # US-NR1: half-hourly = 1800 s
timestep_mms <- 3600L   # US-MMS: hourly = 3600 s

# ── Random seed ────────────────────────────────────────────────────────────────
set.seed(2026L)

# ── Initial conditions ────────────────────────────────────────────────────────
X.orig <- matrix(NA_real_, nrow = ne, ncol = 3)
X.orig[, 1] <- 3.0    # Bleaf (Mg C ha⁻¹)
X.orig[, 2] <- 100.0  # Bwood (Mg C ha⁻¹)
X.orig[, 3] <- 100.0  # BSOM  (Mg C ha⁻¹)

# ── ENF parameters for US-NR1 ─────────────────────────────────────────────────
steps_per_yr_nr1 <- 365L * 48L   # 17,520 half-hourly steps per non-leap year
lf_annual_enf    <- 0.25          # 25% annual leaf turnover (4-year needle life)
mort_annual      <- 0.0125        # 1.25% woody biomass turnover per year

params_nr1 <- data.frame(
  alpha      = rnorm(ne, mean = 0.02,  sd = 0.003) |> pmax(0),
  SLA        = rnorm(ne, mean = 8.0,   sd = 1.2)   |> pmax(1),
  Rbasal     = rnorm(ne, mean = 0.015, sd = 0.003)  |> pmax(0),
  Q10        = rnorm(ne, mean = 2.0,   sd = 0.3)   |> pmax(1),
  litterfall = rnorm(ne,
                     mean = lf_annual_enf / steps_per_yr_nr1,
                     sd   = lf_annual_enf * 0.1 / steps_per_yr_nr1) |> pmax(0),
  mortality  = rnorm(ne,
                     mean = mort_annual / steps_per_yr_nr1,
                     sd   = mort_annual * 0.1 / steps_per_yr_nr1) |> pmax(0),
  sigma.leaf = rep(0.01, ne),
  sigma.stem = rep(0.10, ne),
  sigma.soil = rep(0.10, ne)
)
falloc_nr1 <- rdirichlet.orig(ne, alpha = c(5, 3, 2))  # ENF: Ra 50%, wood 30%, leaf 20%
params_nr1$falloc.1 <- falloc_nr1[, 1]
params_nr1$falloc.2 <- falloc_nr1[, 2]
params_nr1$falloc.3 <- falloc_nr1[, 3]
stopifnot(all(abs(rowSums(params_nr1[, c("falloc.1","falloc.2","falloc.3")]) - 1) < 1e-10))

# ── DBF parameters for US-MMS ─────────────────────────────────────────────────
steps_per_yr_mms <- 365L * 24L   # 8,760 hourly steps per non-leap year
lf_annual_dbf    <- 1.0           # 100% annual leaf turnover (deciduous)

params_mms <- data.frame(
  alpha      = rnorm(ne, mean = 0.02,  sd = 0.003) |> pmax(0),
  SLA        = rnorm(ne, mean = 8.0,   sd = 1.2)   |> pmax(1),
  Rbasal     = rnorm(ne, mean = 0.015, sd = 0.003)  |> pmax(0),
  Q10        = rnorm(ne, mean = 2.0,   sd = 0.3)   |> pmax(1),
  litterfall = rnorm(ne,
                     mean = lf_annual_dbf / steps_per_yr_mms,
                     sd   = lf_annual_dbf * 0.1 / steps_per_yr_mms) |> pmax(0),
  mortality  = rnorm(ne,
                     mean = mort_annual / steps_per_yr_mms,
                     sd   = mort_annual * 0.1 / steps_per_yr_mms) |> pmax(0),
  sigma.leaf = rep(0.01, ne),
  sigma.stem = rep(0.10, ne),
  sigma.soil = rep(0.10, ne)
)
falloc_mms <- rdirichlet.orig(ne, alpha = c(4, 3, 3))  # DBF: Ra 40%, wood 30%, leaf 30%
params_mms$falloc.1 <- falloc_mms[, 1]
params_mms$falloc.2 <- falloc_mms[, 2]
params_mms$falloc.3 <- falloc_mms[, 3]
stopifnot(all(abs(rowSums(params_mms[, c("falloc.1","falloc.2","falloc.3")]) - 1) < 1e-10))

cat(sprintf("params_nr1: %d members × %d parameters (ENF)\n", nrow(params_nr1), ncol(params_nr1)))
cat(sprintf("params_mms: %d members × %d parameters (DBF)\n", nrow(params_mms), ncol(params_mms)))
```

---

## Run prior ensembles

```{r model_run, cache=TRUE}
# ── US-NR1 prior ensemble ──────────────────────────────────────────────────────
message("Starting US-NR1 prior ensemble (expect ~2–4 minutes for 17,568 steps) …")
output_nr1 <- ensemble_forecast(
  X      = X.orig,
  params = params_nr1,
  inputs = as.data.frame(inputs_nr1[, c("PAR", "temp")])
)
cat(sprintf("\noutput_nr1 dimensions: %s\n",
            paste(dim(output_nr1), collapse = " × ")))

# ── US-MMS prior ensemble ──────────────────────────────────────────────────────
message("Starting US-MMS prior ensemble (expect ~1–2 minutes for 8,760 steps) …")
output_mms <- ensemble_forecast(
  X      = X.orig,
  params = params_mms,
  inputs = as.data.frame(inputs_mms[, c("PAR", "temp")])
)
cat(sprintf("\noutput_mms dimensions: %s\n",
            paste(dim(output_mms), collapse = " × ")))
```

---

## Observation vectors

```{r obs_vectors}
# ── Build NEP observation vectors for each site ────────────────────────────────
# SSEM output[,,6] is NEP (positive = C sink; ecological convention).
# FLUXNET NEE > 0 = atmospheric source.  NEP = -NEE.

nep_nr1    <- -flux_nr1$NEE_VUT_REF
nep.qc_nr1 <- as.numeric(flux_nr1$NEE_VUT_REF_QC)
qaqc_nr1   <- nep.qc_nr1 == 0   # TRUE = directly measured

nep_mms    <- -flux_mms$NEE_VUT_REF
nep.qc_mms <- as.numeric(flux_mms$NEE_VUT_REF_QC)
qaqc_mms   <- nep.qc_mms == 0

cat(sprintf("US-NR1 nep: length %d | NA count: %d | measured (QC=0): %d (%.1f%%)\n",
            length(nep_nr1), sum(is.na(nep_nr1)),
            sum(qaqc_nr1, na.rm=TRUE), 100*mean(qaqc_nr1, na.rm=TRUE)))
cat(sprintf("US-MMS nep: length %d | NA count: %d | measured (QC=0): %d (%.1f%%)\n",
            length(nep_mms), sum(is.na(nep_mms)),
            sum(qaqc_mms, na.rm=TRUE), 100*mean(qaqc_mms, na.rm=TRUE)))
```

---

## Save workspaces

```{r save_workspace}
output_dir <- "output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# ── Save US-NR1 workspace ──────────────────────────────────────────────────────
workspace_nr1 <- list(
  ne              = ne,
  timestep        = timestep_nr1,
  inputs          = inputs_nr1,
  X.orig          = X.orig,
  params          = params_nr1,
  output.ensemble = output_nr1,
  flux            = flux_nr1,
  nep             = nep_nr1,
  nep.qc          = nep.qc_nr1,
  qaqc            = qaqc_nr1
)
saveRDS(workspace_nr1, file.path(data_dir, "handoff_workspace_NR1.RDS"))
cat(sprintf("Workspace saved: %s  (%.1f MB)\n",
            file.path(data_dir, "handoff_workspace_NR1.RDS"),
            file.size(file.path(data_dir, "handoff_workspace_NR1.RDS")) / 1e6))

# ── Save US-MMS workspace ──────────────────────────────────────────────────────
workspace_mms <- list(
  ne              = ne,
  timestep        = timestep_mms,
  inputs          = inputs_mms,
  X.orig          = X.orig,
  params          = params_mms,
  output.ensemble = output_mms,
  flux            = flux_mms,
  nep             = nep_mms,
  nep.qc          = nep.qc_mms,
  qaqc            = qaqc_mms
)
saveRDS(workspace_mms, file.path(data_dir, "handoff_workspace_MMS.RDS"))
cat(sprintf("Workspace saved: %s  (%.1f MB)\n",
            file.path(data_dir, "handoff_workspace_MMS.RDS"),
            file.size(file.path(data_dir, "handoff_workspace_MMS.RDS")) / 1e6))

# ── Also write to the legacy output/ path that earlier exercise versions used ─
# Mike's PM code may load from exercises/output/handoff_workspace.RDS.
# Write the selected handoff_site workspace to that path as a convenience copy.
legacy_path <- file.path(output_dir, "handoff_workspace.RDS")
selected_ws <- if (handoff_site == "US-NR1") workspace_nr1 else workspace_mms
saveRDS(selected_ws, legacy_path)
cat(sprintf("Legacy path:    %s  (copy of %s workspace)\n",
            legacy_path, handoff_site))
```

---

## Checklist

```{r checklist}
# ── Reload and verify both workspaces ─────────────────────────────────────────
check_workspace <- function(ws_path, site_label, n_ts_expect, ne_expect, params_expect) {
  ws <- readRDS(ws_path)

  checks <- list(
    list(name = "ne",              expect_class = "integer",    expect_dim = paste0("scalar ", ne_expect)),
    list(name = "timestep",        expect_class = "integer",    expect_dim = paste0("scalar ", ws$timestep)),
    list(name = "inputs",          expect_class = "data.frame", expect_dim = paste0(n_ts_expect, " × 3")),
    list(name = "X.orig",          expect_class = "matrix",     expect_dim = paste0(ne_expect, " × 3")),
    list(name = "params",          expect_class = "data.frame", expect_dim = paste0(ne_expect, " × ", params_expect)),
    list(name = "output.ensemble", expect_class = "array",
         expect_dim = paste0(n_ts_expect, " × ", ne_expect, " × 12")),
    list(name = "flux",            expect_class = "data.frame", expect_dim = paste0(n_ts_expect, " × 6")),
    list(name = "nep",             expect_class = "numeric",    expect_dim = paste0("length ", n_ts_expect)),
    list(name = "nep.qc",          expect_class = "numeric",    expect_dim = paste0("length ", n_ts_expect)),
    list(name = "qaqc",            expect_class = "logical",    expect_dim = paste0("length ", n_ts_expect))
  )

  cat("\n")
  cat(strrep("=", 72), "\n")
  cat(sprintf("CHECKLIST — %s  (%s)\n", site_label, basename(ws_path)))
  cat(strrep("=", 72), "\n")
  cat(sprintf("  %-20s  %-12s  %-22s  %s\n",
              "Object", "Class", "Dimensions", "Status"))
  cat(strrep("-", 72), "\n")

  all_ok <- TRUE
  for (chk in checks) {
    obj <- ws[[chk$name]]
    # Compute actual class and dim
    actual_class <- class(obj)[1]
    actual_dim <- if (is.null(dim(obj))) {
      if (length(obj) == 1L) paste0("scalar ", obj)
      else                   paste0("length ", length(obj))
    } else paste(dim(obj), collapse = " × ")

    class_ok <- switch(chk$expect_class,
      "data.frame" = inherits(obj, "data.frame"),
      "numeric"    = is.numeric(obj),
      class(obj)[1] == chk$expect_class
    )
    dim_ok <- (actual_dim == chk$expect_dim)
    ok     <- class_ok && dim_ok
    if (!ok) all_ok <- FALSE
    status <- if (ok) "[OK]" else "[FAIL]"
    cat(sprintf("  %-20s  %-12s  %-22s  %s\n",
                chk$name, actual_class, actual_dim, status))
    if (!class_ok) cat(sprintf("    expected class: %s\n", chk$expect_class))
    if (!dim_ok)   cat(sprintf("    expected dim:   %s\n", chk$expect_dim))
  }
  cat(strrep("-", 72), "\n")
  if (all_ok) {
    cat(sprintf("All objects verified.  %s workspace is ready.\n", site_label))
  } else {
    warning(sprintf("One or more objects failed for %s — see [FAIL] rows above.",
                    site_label))
  }

  # Spot checks
  nep_obs_mean <- mean(ws$nep[ws$qaqc], na.rm = TRUE)
  nep_mod_mean <- mean(ws$output.ensemble[ws$qaqc, , 6], na.rm = TRUE)
  cat(sprintf("\nSpot checks:\n"))
  cat(sprintf("  Observed NEP mean at QC=0: %+.3f umol m-2 s-1\n", nep_obs_mean))
  cat(sprintf("  Prior SSEM NEP mean at QC=0: %+.3f umol m-2 s-1\n", nep_mod_mean))
  cat(sprintf("  inputs date range: %s  to  %s\n",
              format(min(ws$inputs$date), "%Y-%m-%d"),
              format(max(ws$inputs$date), "%Y-%m-%d")))
  cat(strrep("=", 72), "\n")

  invisible(all_ok)
}

# US-NR1: 2008 (leap year) × 48 half-hourly steps = 17,568 timesteps.
ok_nr1 <- check_workspace(
  file.path(data_dir, "handoff_workspace_NR1.RDS"),
  "US-NR1 (ENF, 2008)",
  n_ts_expect    = 17568L,
  ne_expect      = ne,
  params_expect  = 12L
)

# US-MMS: 2005 (non-leap year) × 24 hourly steps = 8,760 timesteps
ok_mms <- check_workspace(
  file.path(data_dir, "handoff_workspace_MMS.RDS"),
  "US-MMS (DBF, 2005)",
  n_ts_expect    = 8760L,
  ne_expect      = ne,
  params_expect  = 12L
)

# ── Final summary ──────────────────────────────────────────────────────────────
cat("\n")
cat(strrep("=", 72), "\n")
cat("HANDOFF SUMMARY\n")
cat(strrep("=", 72), "\n")
cat(sprintf("  handoff_site setting: %s\n", handoff_site))
cat(sprintf("  Selected site workspace (legacy path):\n"))
cat(sprintf("    exercises/output/handoff_workspace.RDS\n\n"))
cat(sprintf("  US-NR1 workspace: %s\n",
            file.path(data_dir, "handoff_workspace_NR1.RDS")))
cat(sprintf("  US-MMS workspace: %s\n",
            file.path(data_dir, "handoff_workspace_MMS.RDS")))
cat(strrep("-", 72), "\n")

if (ok_nr1 && ok_mms) {
  cat("Both workspaces verified.  Raise your hand — PM session can start.\n")
  cat("Mike's code will load the workspace with:\n")
  cat(sprintf("  ws <- readRDS('%s')\n",
              file.path(data_dir, sprintf("handoff_workspace_%s.RDS",
                                          sub("US-", "", handoff_site)))))
} else {
  stop("One or more workspace checks failed — see [FAIL] rows above.")
}
cat(strrep("=", 72), "\n")
```
